---
title: "reproducing_kim_BASE47"
format: html
editor: visual
---

This is running BASE47 on **their data** and trying to reproduce their results

# Workflow

bulk RNA-seq per-sample expression matrix

↓

normalized/log-transformed expression

↓

subset to BASE47 genes only

↓

RNA-seq BASE47 classifier

↓

Basal vs Luminal call per sample

# **Setup**

```{r}
#| message: false
#| warning: false
#| code-fold: true

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

if (!requireNamespace("DESeq2", quietly = TRUE)) {
  BiocManager::install("DESeq2")
}

if (!requireNamespace("biomaRt", quietly = TRUE)) {
  BiocManager::install("biomaRt")
}

if (!requireNamespace("pamr", quietly = TRUE)) {
  install.packages("pamr")
}

library(here)
library(DESeq2)
library(pamr)
library(tidyverse)
```

# **BASE47 Genes**

```{r}
base47_genes <- c(
  "ALOX5AP", "BHMT", "CAPN5", "CD14", "CDK6", "CYP2J2", "CYP4B1",
  "EMP3", "FAP", "FBP1", "GATA3", "HMGCS2", "MSN", "MT1X", "MT2A",
  "PRRX1", "PPARG", "SCNN1B", "SCNN1G", "SLC9A2", "SPINK1", "TBX2",
  "TMPRSS2", "UPK2", "PPFIBP2", "SEMA5A", "ADIRF", "SLC27A2",
  "GLIPR1", "UPK1A", "TRAK1", "PALLD", "GPD1L", "TOX3", "TMEM97",
  "CHST15", "VGLL1", "PLEKHG6", "PDGFC", "GAREM1", "GDPD3", "RNF128",
  "TUBB6", "CAVIN3", "AHNAK2", "RAB15", "FAM174B"
)
```

# **RNAseq Input**

- Data was obtained from GEO accession GSE160693 using the normalized gene count matrix file `GSE160693_Normalized_Gene_Counts_Matrix.txt.gz`

```{r}
#########################
# Load and format Kim Lab RNA-seq data
#########################
kim_data <- read.delim(
  here("data", "GSE160693_Normalized_Gene_Counts_Matrix.txt"),
  row.names = 1,
  check.names = FALSE
)

# format and keep only gene symbols
gene_symbols <- sub("\\|.*$", "", rownames(kim_data))
kim_data <- kim_data[!duplicated(gene_symbols), ]      # dedupe rows first
gene_symbols <- gene_symbols[!duplicated(gene_symbols)]  # matching dedupe on the vector
rownames(kim_data) <- gene_symbols                 # now safe to assign


#########################
# Match BASE47 gene names
#########################

#alt names
alias_map <- c(ADIRF = "C10orf116", CAVIN3 = "PTRF", GAREM1 = "GAREM")

geo_lookup_id <- ifelse(base47_genes %in% names(alias_map),
                        alias_map[base47_genes], base47_genes)

# subset rnaseq to only base47 genes (check which BASE47 genes are present in data)
present <- geo_lookup_id %in% rownames(kim_data)
if (any(!present)) {
  warning("Missing from your data: ", paste(base47_genes[!present], collapse = ", "))
}

#########################
# Prepare BASE47 test matrix
#########################

test_matrix <- as.matrix(kim_data[geo_lookup_id[present], , drop = FALSE])
rownames(test_matrix) <- base47_genes[present]

#median-center each gene
test_input <- sweep(test_matrix, 1, apply(test_matrix, 1, median, na.rm = TRUE), "-")

# add 0 rows for any missing BASE47 genes
missing_genes <- setdiff(base47_genes, rownames(test_input))
if (length(missing_genes) > 0) {
  zero_rows <- matrix(0, nrow = length(missing_genes), ncol = ncol(test_input),
                      dimnames = list(missing_genes, colnames(test_input)))
  test_input <- rbind(test_input, zero_rows)
}
# put genes in EXACT BASE47 order
test_input <- test_input[base47_genes, ]
# confirm genes are median-centered
stopifnot(all(abs(apply(test_input, 1, median, na.rm = TRUE)) < 1e-10))

#save input to BASE47 before running
write.table(
  test_input,
  here("data", "kimlab_BASE47_INPUT.txt"),
  sep = "\t",
  quote = FALSE,
  col.names = NA
)

#########################
# Load and prepare BASE47 training data
#########################

# run base47
x <- read.table(here("data","BASE47_RNASEQ_PREDICTOR_DO_NOT_EDIT.txt"),sep="\t",row.names=1,header=T)
classes <- as.vector(t(x[1,]))
xn <- apply(x[-1,],2,function(x){as.numeric(as.vector(x))})
rownames(xn) <- rownames(x)[-1]

# update gene names
rownames(xn)[rownames(xn) == "GAREM"]   <- "GAREM1"
rownames(xn)[rownames(xn) == "PRKCDBP"] <- "CAVIN3"
xn <- xn[base47_genes, ] #make sure same gene order as BASE47

#########################
# Run BASE47 classifier
#########################

test <- read.table(here("data","kimlab_BASE47_INPUT.txt"), sep="\t",row.names=1,header=T)

trainSet <- list(x=scale(xn), y=classes, geneid=rownames(xn), genenames=rownames(xn))
mytrain <-pamr.train(trainSet)
pred.class <- pamr.predict(mytrain,scale(test),0)
pred.prob <- pamr.predict(mytrain,scale(test),0,type="posterior")

#########################
# Format BASE47 predictions
#########################
unc.tma.rna.calls_kim <- as.data.frame(cbind(pred.prob,pred.class))
colnames(unc.tma.rna.calls_kim) <- c("Basal.Correlation","Luminal.Correlation","BASE47.Subtype")
unc.tma.rna.calls_kim$BASE47.Subtype[unc.tma.rna.calls_kim$BASE47.Subtype == 1] <- "Basal"
unc.tma.rna.calls_kim$BASE47.Subtype[unc.tma.rna.calls_kim$BASE47.Subtype == 2] <- "Luminal"
unc.tma.rna.calls_kim

unc.tma.rna.calls_kim_formatted <- unc.tma.rna.calls_kim %>%
  rownames_to_column("Sample") %>%
  mutate(
    Tumor.Model = case_when(
      str_detect(Sample, "^BBN") ~ "BBN",
      str_detect(Sample, "^UPPL") ~ "UPPL",
      TRUE ~ NA_character_
    ),
    Basal.Score = Basal.Correlation - Luminal.Correlation
  ) %>%
  arrange(Basal.Score)

#extra formatting sort alphabetically by sample name
unc.tma.rna.calls_kim_formatted <- unc.tma.rna.calls_kim_formatted %>%
  mutate(
    Sample = factor(
      Sample,
      levels = unique(Sample[order(
        sub("[0-9]+$", "", Sample),
        as.numeric(sub(".*?([0-9]+)$", "\\1", Sample))
      )])
    )
  )

#########################
# Confusion matrix
#########################

# 1. Luminal samples together
# 2. Basal samples together
# 3. within each group, sort by Basal.Score
confusion_mtx <- table(
  Expected = unc.tma.rna.calls_kim_formatted$Tumor.Model,
  Called = unc.tma.rna.calls_kim_formatted$BASE47.Subtype
)
confusion_mtx <- as.data.frame.matrix(confusion_mtx)

#########################
# Waterfall plot
#########################
wfall_basal_score <- ggplot(
  plot_data,
  aes(
    x = Sample,
    y = Basal.Score,
    fill = BASE47.Subtype
  )
) +
  geom_col(width = 0.8) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.5,
    color = "black"
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "BASE47 score (Luminal ← 0 → Basal)",
    fill = "BASE47 Subtype"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 8),
    legend.position = "top"
  )

ggsave(
  filename = here("outs", "kimlab_BASE47_reproduced_waterfall.png"),
  plot = wfall_basal_score,
  width = 7,
  height = 6,
  dpi = 300
)

wfall_basal_score
```
