---
title: "run_BASE47"
format:
  html:
    code-tools: true
    code-link: true
editor: visual
---

This is running BASE47 on **our data**

# Workflow

bulk RNA-seq per-sample expression matrix

↓

normalized/log-transformed expression

↓

subset to BASE47 genes only

↓

RNA-seq BASE47 classifier

↓

Basal vs Luminal call per sample

# **Setup**

- Every output gets saved to /**outs**

```{r}
#| message: false
#| warning: false
#| code-fold: true

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

if (!requireNamespace("DESeq2", quietly = TRUE)) {
  BiocManager::install("DESeq2")
}

if (!requireNamespace("biomaRt", quietly = TRUE)) {
  BiocManager::install("biomaRt")
}

if (!requireNamespace("pamr", quietly = TRUE)) {
  install.packages("pamr")
}

library(here)
library(DESeq2)
library(pamr)
library(tidyverse)
```

# **BASE47 Genes**

- These are already predefined by paper

- **ADIRF** isn't in our data so it's omitted (now BASE46 technically)

```{r}
base47_genes <- c(
  "ALOX5AP", "BHMT", "CAPN5", "CD14", "CDK6", "CYP2J2", "CYP4B1",
  "EMP3", "FAP", "FBP1", "GATA3", "HMGCS2", "MSN", "MT1X", "MT2A",
  "PRRX1", "PPARG", "SCNN1B", "SCNN1G", "SLC9A2", "SPINK1", "TBX2",
  "TMPRSS2", "UPK2", "PPFIBP2", "SEMA5A", "ADIRF", "SLC27A2",
  "GLIPR1", "UPK1A", "TRAK1", "PALLD", "GPD1L", "TOX3", "TMEM97",
  "CHST15", "VGLL1", "PLEKHG6", "PDGFC", "GAREM1", "GDPD3", "RNF128",
  "TUBB6", "CAVIN3", "AHNAK2", "RAB15", "FAM174B"
)
```

# **RNAseq Input**

- The input data for this analysis is an RNA-seq expression matrix generated from Plasmidsaurus sequencing data

- **Dataset**

  - **Comparison:** Parental vs. MP1

  - **Plasmidsaurus ID:** K8HVZP

  - **Sequencing date:** May 18, 2026

  - **Original data format:** Expression matrix (`.tsv`)

  - **File used for analysis:** `parental_mp1_K8HVZP-expression-matrix.tsv`

  - **Data type:** CPM-normalized expression counts

- Raw data for BBN+UPPL is cpm but BASE47 requires 90th percentile, log2(tpm+1)

  - So I convert cpm to log2(cpm+1)

```{r}
#expression matrix load in
RNAseq_raw <- read.table(here("data","parental_mp1_K8HVZP-expression-matrix_cpm_counts.csv"), sep=",", row.names=1, header=T)

# Filter to either only MP1, only parentals, only BBNs, only UPPLs
# RNAseq_raw <- RNAseq_raw %>% select(contains("MP1"))
# RNAseq_raw <- RNAseq_raw %>% select(contains("parental"))
# RNAseq_raw <- RNAseq_raw %>% select(contains("BBN"))
# RNAseq_raw <- RNAseq_raw %>% select(contains("UPPL"))

#Drop protein-coding identity + raw counts columns // keep only cpm columns
RNAseq_raw <- RNAseq_raw[, !grepl("count|^gene_biotype", colnames(RNAseq_raw)    )    ]

# mouse ensembl id to human gene id mapping (MANUALLY, b/c bioMarT was down)
base47_mouse_map <- read.table(
  text = "
ENSMUSG00000000093	TBX2
ENSMUSG00000000216	SCNN1G
ENSMUSG00000000385	TMPRSS2
ENSMUSG00000000392	FAP
ENSMUSG00000000440	PPARG
ENSMUSG00000001473	TUBB6
ENSMUSG00000006313	UPK1A
ENSMUSG00000015619	GATA3
ENSMUSG00000021062	RAB15
ENSMUSG00000022231	SEMA5A
ENSMUSG00000024503	SPINK1
ENSMUSG00000026062	SLC9A2
ENSMUSG00000026586	PRRX1
ENSMUSG00000027359	SLC27A2
ENSMUSG00000027875	HMGCS2
ENSMUSG00000028019	PDGFC
ENSMUSG00000028713	CYP4B1
ENSMUSG00000030703	GDPD3
ENSMUSG00000030873	SCNN1B
ENSMUSG00000030930	CHST15
ENSMUSG00000031131	VGLL1
ENSMUSG00000031207	MSN
ENSMUSG00000031438	RNF128
ENSMUSG00000032536	TRAK1
ENSMUSG00000035547	CAPN5
ENSMUSG00000036528	PPFIBP2
ENSMUSG00000037060	PRKCDBP
ENSMUSG00000037278	TMEM97
ENSMUSG00000038167	PLEKHG6
ENSMUSG00000040212	EMP3
ENSMUSG00000040274	CDK6
ENSMUSG00000041523	UPK2
ENSMUSG00000042680	GAREM
ENSMUSG00000043668	TOX3
ENSMUSG00000050627	GPD1L
ENSMUSG00000051439	CD14
ENSMUSG00000056888	GLIPR1
ENSMUSG00000058056	PALLD
ENSMUSG00000060063	ALOX5AP
ENSMUSG00000069805	FBP1
ENSMUSG00000074768	BHMT
ENSMUSG00000078670	FAM174B
ENSMUSG00000118667	AHNAK2
ENSMUSG00000031762	MT2A
ENSMUSG00000031765	MT1X
ENSMUSG00000052914	CYP2J2
NONE	ADIRF
",
  sep = "\t",
  col.names = c("mouse_ensembl", "human_gene"),
  stringsAsFactors = FALSE
)

# Convert mouse ensembl ID to human orthologs
RNAseq_raw_base47 <- RNAseq_raw %>%
  as.data.frame() %>%
  rownames_to_column("mouse_ensembl") %>%
  dplyr::mutate(mouse_ensembl_clean = sub("\\..*$", "", mouse_ensembl)) %>%
  #filter RNAseq input to only BASE47 genes
  dplyr::inner_join(                    
    base47_mouse_map %>%
      dplyr::filter(mouse_ensembl != "NONE") %>%
      dplyr::select(mouse_ensembl, human_gene),
    by = c("mouse_ensembl_clean" = "mouse_ensembl")
  ) %>%
  dplyr::select(-dplyr::any_of(c("mouse_ensembl", "mouse_ensembl_clean", "gene_type"))) %>%
  column_to_rownames("human_gene")

RNAseq_raw_base47$gene_name <- NULL

# mouse genes that didn't have human ortholog
missing_human_genes <- setdiff(
  base47_mouse_map$human_gene,
  rownames(RNAseq_raw_base47)
)
if (length(missing_human_genes) > 0) {
  cat(
    "Genes that didn't have human ortholog:",
    paste(missing_human_genes, collapse = ", "),
    "\n"
  )
}

# Ensure matrix is numeric
RNAseq_raw_base47[] <- lapply(RNAseq_raw_base47, as.numeric)

# Convert cpm to log2 upper quartile-normalized expression (that's how they did it in their methods)
#   1. For each sample, calculate 75th percentile of nonzero CPM values
#   2. Divide each sample by its own upper-quartile value
#   3. Multiply by 1000 to put values on comparable scale
#   4. Apply log2(x+1) transformation
cpm_to_uq_log2 <- function(cpm_matrix) {
  uq <- apply(cpm_matrix, 2, function(x) quantile(x[x > 0], 0.75))
  scaled <- sweep(cpm_matrix, 2, uq, "/") * 1000
  log2(scaled + 1)
}

base47_uqlog2 <- RNAseq_raw_base47 %>% 
  cpm_to_uq_log2() %>% 
  rename_with(~ str_remove(., "_cpm"), ends_with("_cpm")) 

# median-center every gene by subtracting its row median from every sample (median-centers)
base47_input <- sweep(base47_uqlog2, 1, apply(base47_uqlog2, 1, median, na.rm = TRUE), "-")

# Every row median should now be exactly 0 (sanity check that matrix is converted correctly)
stopifnot(all(abs(apply(base47_input, 1, median, na.rm = TRUE)) < 1e-10))

#remove variables to save memory
# rm(base47_mouse_map, RNAseq_raw, RNAseq_raw_base47)
```

# **Load predictor**

```{r}
# Load BASE47 RNA-seq prediction matrix
download.file(
  "https://raw.githubusercontent.com/kimlabunc/Kardos_Rose_NanoString/main/BASE47_RNASEQ_PREDICTOR_DO_NOT_EDIT.txt",
  destfile = here("data", "BASE47_RNASEQ_PREDICTOR_DO_NOT_EDIT.txt")
)

# ==================
# x = RNAseq training/predictor matrix
# ==================
x <- read.table(
  here("data", "BASE47_RNASEQ_PREDICTOR_DO_NOT_EDIT.txt"),
  sep = "\t",
  row.names = 1,
  header = TRUE
)

# gets labels for subtypes from 1ST ROW where: 1 = Basal; 2=Luminal
classes <- as.vector(t(x[1,]))

# removes 1st row, converts expression values to numeric, and restores gene names as row names
#     b/c xn expects: ONLY gene x samples, so you can't have extraneous columns/rows
xn <- apply(x[-1,],2,function(x){as.numeric(as.vector(x))}) # xn is training expression matrix: genes by samples
rownames(xn) <- rownames(x)[-1]

missing_genes <- setdiff(rownames(xn), rownames(base47_input))

#Checks for missing BASE47 genes in our data
#     it's ok if it stops the script, it's just mainly to let you know which ones are missing
#     if missing any genes, then run the BASE47 classifier on only the genes we have, rather than making the expression=0 for that missing gene

# if (length(missing_genes) > 0) {
#   stop("Missing BASE47 genes from df: ", paste(missing_genes, collapse = ", "))
# }

# if missing any genes, remove it from prediction matrix
# rather than making expression=0, you will instead classify using less than 47 genes instead
common_genes <- intersect(rownames(base47_uqlog2), rownames(xn))

# Optional: see what was removed
missing_from_sample <- setdiff(rownames(xn), rownames(base47_uqlog2))
if (length(missing_from_sample) > 0) {
  cat("Missing from RNAseq input:", paste(missing_from_sample, collapse = ", "), "\n")
}

# Subset both matrices to the same genes, same order
base47_uqlog2_matched <- base47_uqlog2[common_genes, , drop = FALSE]
xn_matched <- xn[common_genes, , drop = FALSE]
```

# **Molecular Subtype Classifier**

```{r}
# Classify BBN963+UPPL1541 RNAseq file
df <- base47_input[rownames(xn_matched), , drop = FALSE]

# training lines taken from kim lab repo VVV
trainSet <- list(x=scale(xn_matched), y=classes, geneid=rownames(xn_matched), genenames=rownames(xn_matched))
mytrain <-pamr.train(trainSet)
pred.class <- pamr.predict(mytrain,scale(df),0)
pred.prob <- pamr.predict(mytrain,scale(df),0,type="posterior")
unc.tma.rna.calls <- as.data.frame(cbind(pred.prob,pred.class))
colnames(unc.tma.rna.calls) <- c("Basal.Correlation","Luminal.Correlation","BASE47.Subtype")
unc.tma.rna.calls$BASE47.Subtype[unc.tma.rna.calls$BASE47.Subtype == 1] <- "Basal"
unc.tma.rna.calls$BASE47.Subtype[unc.tma.rna.calls$BASE47.Subtype == 2] <- "Luminal"

write.csv(unc.tma.rna.calls, file="BASE47_classification_output.csv", row.names=TRUE)
```

# **Waterfall Classification Plot**

```{r}
#Create new columns for expected mol subtype and its BASE47 call
unc.tma.rna.calls_formatted <- unc.tma.rna.calls %>%
  rownames_to_column("Sample") %>%
  mutate(
    Tumor.Model = case_when(
      str_detect(Sample, "^BBN") ~ "BBN",
      str_detect(Sample, "^UPPL") ~ "UPPL",
      TRUE ~ NA_character_
    ),
    Basal.Score = Basal.Correlation - Luminal.Correlation
  ) %>%
  arrange(Basal.Score)

#extra formatting sort alphabetically by sample name
unc.tma.rna.calls_formatted <- unc.tma.rna.calls_formatted %>%
  mutate(
    Sample = factor(
      Sample,
      levels = unique(Sample[order(
        sub("[0-9]+$", "", Sample),
        as.numeric(sub(".*?([0-9]+)$", "\\1", Sample))
      )])
    )
  )

# Confusion matrix
confusion_mtx <- table(
  Expected = unc.tma.rna.calls_formatted$Tumor.Model,
  Called = unc.tma.rna.calls_formatted$BASE47.Subtype
)
confusion_mtx <- as.data.frame.matrix(confusion_mtx)
write.csv(confusion_mtx, 
          file=here("outs","confusion_matrix.csv"),
          row.names=TRUE)
confusion_mtx

# Waterfall plot
wfall_basal_score <- ggplot(
  unc.tma.rna.calls_formatted,
  aes(
    x = Sample,
    y = Basal.Score,
    color = Tumor.Model,
    fill = Tumor.Model
  )
) +
  geom_bar(stat = "identity") +
  scale_fill_hue(c = 40) +
  coord_flip() +
  labs(
    title = "BASE47 Molecular Subtype Classification\nof BBN963 and UPPL1541 Samples",
    x = "Sample",
    y = "Basal centroid score: Luminal = -1, Basal = +1",
    fill = "Tumor Model",
    color = "Tumor Model"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    )
  )

ggsave(here("outs","BASE47_molecular_subtype_classification.png"), wfall_basal_score, width = 7, height = 5, dpi = 400)

wfall_basal_score
```

```{r}
threshold <- 0

gene_scores <- pamr.listgenes(
  fit = mytrain,
  data = trainSet,
  threshold = threshold,
  genenames = TRUE
)

gene_scores <- data.frame(
  gene_scores,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

names(gene_scores)[names(gene_scores) == "1-score"] <-
  "Basal.Centroid.Score"

names(gene_scores)[names(gene_scores) == "2-score"] <-
  "Luminal.Centroid.Score"

gene_scores <- gene_scores %>%
  mutate(
    Basal.Centroid.Score = as.numeric(Basal.Centroid.Score),
    Luminal.Centroid.Score = as.numeric(Luminal.Centroid.Score),
    Basal.Direction =
      Basal.Centroid.Score - Luminal.Centroid.Score,
    Marker.Type = case_when(
      Basal.Direction > 0 ~ "Basal",
      Basal.Direction < 0 ~ "Luminal",
      TRUE ~ "Neutral"
    )
  ) %>%
  arrange(desc(abs(Basal.Direction)))

write.csv(
  gene_scores,
  "BASE47_contributing_gene_centroids.csv",
  row.names = FALSE
)

gene_scores
```

# Heatmap Plot of Gene Contributions

```{r}

#data format
df_scaled <- scale(df)
basal_level <- "1"
luminal_level <- "2"

# Shrunken class centroids at chosen threshold
centroid_matrix <- pamr.predict(
  mytrain,
  df_scaled,
  threshold = threshold,
  type = "centroid"
)

# Standardized centroid deviations
delta <- sweep(
  centroid_matrix,
  1,
  mytrain$centroid.overall,
  "-"
)

delta <- sweep(
  delta,
  1,
  mytrain$sd,
  "/"
)

# Standardized test expression
x_standardized <- sweep(
  df_scaled,
  1,
  mytrain$centroid.overall,
  "-"
)

x_standardized <- sweep(
  x_standardized,
  1,
  mytrain$sd,
  "/"
)

# Genes surviving the threshold
nonzero_index <- pamr.predict(
  mytrain,
  df_scaled,
  threshold = threshold,
  type = "nonzero"
)

delta_difference <-
  delta[nonzero_index, basal_level] -
  delta[nonzero_index, luminal_level]

delta_squared_difference <-
  delta[nonzero_index, basal_level]^2 -
  delta[nonzero_index, luminal_level]^2

gene_contribution <- sweep(
  x_standardized[nonzero_index, , drop = FALSE],
  1,
  delta_difference,
  "*"
)

gene_contribution <- sweep(
  gene_contribution,
  1,
  0.5 * delta_squared_difference,
  "-"
)

contribution_long <- data.frame(
  gene_contribution,
  check.names = FALSE
) %>%
  rownames_to_column("Gene") %>%
  pivot_longer(
    -Gene,
    names_to = "Sample",
    values_to = "Contribution"
  )

expression_long <- data.frame(
  df_scaled[nonzero_index, , drop = FALSE],
  check.names = FALSE
) %>%
  rownames_to_column("Gene") %>%
  pivot_longer(
    -Gene,
    names_to = "Sample",
    values_to = "Scaled.Expression"
  )

contribution_long <- contribution_long %>%
  left_join(
    expression_long,
    by = c("Gene", "Sample")
  ) %>%
  mutate(
    Supports = if_else(
      Contribution >= 0,
      "Basal",
      "Luminal"
    )
  )

top_contributors <- contribution_long %>%
  arrange(Sample, desc(abs(Contribution))) %>%
  group_by(Sample) %>%
  slice_head(n = 10) %>%
  ungroup()

write.csv(
  contribution_long, here("outs",
  "BASE47_gene_contributions_by_sample.csv"),
  row.names = FALSE
)

top_contributors

top_genes <- contribution_long %>%
  group_by(Gene) %>%
  summarise(
    Maximum.Absolute.Contribution =
      max(abs(Contribution)),
    .groups = "drop"
  ) %>%
  slice_max(
    Maximum.Absolute.Contribution,
    n = 25,
    with_ties = FALSE
  ) %>%
  pull(Gene)

contribution_plot_data <- contribution_long %>%
  filter(Gene %in% top_genes) %>%
  mutate(
    Gene = factor(Gene, levels = rev(top_genes))
  )

gene_contribution_heatmap <- ggplot(
  contribution_plot_data,
  aes(
    x = Sample,
    y = Gene,
    fill = Contribution
  )
) +
  geom_tile() +
  scale_fill_gradient2(
    midpoint = 0,
    name = "Contribution"
  ) +
  labs(
    title = "Genes Contributing to BASE47 Subtype Calls",
    subtitle = "Positive = Basal; Negative = Luminal",
    x = "Sample",
    y = "Gene"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1
    )
  )

gene_contribution_heatmap
ggsave(
  here("outs","BASE47_gene_contributions_heatmap.png"), plot = gene_contribution_heatmap, dpi=400)
```

Lollipop Plot

- Gene contributions

```{r}
subtype_pal <- c(Basal = "#C0392B", Luminal = "#2C6FA6")

sample_id <- unique(contribution_long$Sample)[1]   # pick which sample to look at
top_n      <- 47

# filter df to get specific sample
lolli_one <- contribution_long %>%
  filter(Sample == sample_id) %>%
  slice_max(abs(Contribution), n = top_n, with_ties = FALSE) %>%
  mutate(Gene = factor(Gene, levels = Gene[order(Contribution)]))  # sort by signed value

# generate lollipop
p_one <- ggplot(lolli_one, aes(x = Contribution, y = Gene, colour = Supports)) +
  geom_segment(aes(x = 0, xend = Contribution, yend = Gene), linewidth = 0.6) +
  geom_vline(xintercept = 0, colour = "grey40", linewidth = 0.4) +
  geom_point(size = 3.2) +
  scale_colour_manual(values = subtype_pal, name = "Supports") +
  labs(
    title    = paste0("Top gene contributions to BASE47 call - ", sample_id),
    subtitle = "Contribution > 0 supports Basal; < 0 supports Luminal",
    x = "Discriminant contribution", y = NULL) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(face = "italic"),   # italic gene symbols
    plot.title         = element_text(face = "plain", size = 11.5),
    plot.subtitle      = element_text(size = 8.5, colour = "grey35"),
    legend.position    = c(0.85, 0.15),
    legend.background   = element_rect(fill = "white", colour = NA)
  )

ggsave(
  here("outs", "BASE47_gene_contributions_lollipop.png"),
  plot = p_one,
  width = 7,
  height = 8,
  dpi = 400
)

p_one
```
