---
title: "GATA3/PPARG Scoring"
format: html
editor: visual
---

Folder structure: GATA3_PPARG_Scoring/ \|-- data/ \|-- results/ \| `-- plots/`-- scripts/ \`-- run_gata3_pparg_files/ ---

# Setup

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(here)
# here::i_am("README.Rmd")

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("GEOquery", "impute"))
library(samr)
library(GEOquery)
library(limma)
library(tidyverse)

# =========
# Functions
# =========

# plotting functions stored separately
source(here::here("scripts", "gata3_pparg_plot_functions.R"))

#Convert to 90th percentile of log(tpm+1)
cpm_to_uq_log2 <- function(cpm_matrix) {
  uq <- apply(cpm_matrix, 2, function(x) quantile(x[x > 0], 0.75))
  scaled <- sweep(cpm_matrix, 2, uq, "/") * 1000
  log2(scaled + 1)
}

# Suppress permutations output
quiet <- function(expr) {
  suppressMessages(suppressWarnings({
    capture.output(result <- eval.parent(substitute(expr)))
    result
  }))
}
```

# Derive Public Rosiglitazone-Induced PPARγ Pathway Signature from GSE47993

1.  Downloads public FABP4-modulation GEO dataset GSE47993
2.  Subset to only UMUC9 control and rosi-treated samples
    1.  Uses rosiglitazone treatment as a PPARγ activation condition
3.  Map Illumina Probe IDs to gene symbols
4.  Collapse multiple probes mapping to the same gene by averaging expression values
5.  Uses SAM to identify genes significantly upregulated after rosi treatment

```{r}
gse47993_dataset <- getGEO("GSE47993", GSEMatrix = TRUE)[[1]]
gse47993_expr <- exprs(gse47993_dataset)

umuc9_rosi_expr <- gse47993_expr[, c("GSM1164283","GSM1164284","GSM1164285",  # control samples
                       "GSM1164286","GSM1164287","GSM1164288") # rosi treated samples
]

# Get gene names from illumina probe IDs (they used IlluminaHT-12 V4.0 expression beadchip)
gse47993_annotation <- fData(gse47993_dataset)
probe_gene_symbols <- gse47993_annotation[rownames(umuc9_rosi_expr), "Symbol"]

# keep probes with valid gene symbols
umuc9_expr_with_symbols <- umuc9_rosi_expr[!is.na(probe_gene_symbols) & probe_gene_symbols != "", ]
probe_gene_symbols <- probe_gene_symbols[!is.na(probe_gene_symbols) & probe_gene_symbols != ""]
# Collapse multiple probes per gene
umuc9_gene_expr <- avereps(umuc9_expr_with_symbols, ID = probe_gene_symbols) # now have gene expression x sample matrix
rosi_group_labels <- c(1,1,1,2,2,2) #1=ctrl, #2=rosi-treated

umuc9_sam_input <- list(
  x = umuc9_gene_expr, 
  y = rosi_group_labels,
  geneid = rownames(umuc9_gene_expr),
  genenames = rownames(umuc9_gene_expr),
  logged2 = TRUE
)

umuc9_sam_fit <- quiet(
  samr(
    data=umuc9_sam_input, resp.type = "Two class unpaired", nperms = 100)
)
umuc9_delta_table <- quiet(
  samr.compute.delta.table(umuc9_sam_fit)
)

# Choose a delta where median FDR < 0.05
# -b/c delta != FDR
sam_delta_results_umuc9 <- as.data.frame(umuc9_delta_table)
fdr05_delta_umuc9 <- sam_delta_results_umuc9[sam_delta_results_umuc9[, "median FDR"] < 0.05, "delta"][1]

# get significant genes in rosi vs ctrl
umuc9_rosi_siggenes <- samr.compute.siggenes.table(
  samr.obj = umuc9_sam_fit,
  del = fdr05_delta_umuc9,
  data = umuc9_sam_input,
  delta.table = umuc9_delta_table
)
```

**Extract Rosi-Induced PPARG Target Genes**

- Saves significantly upregulated rosiglitazone-responsive genes as PPARG activity signature
- These genes represent downstream transcriptional changes associated with PPARG activation

```{r}
# Significant upregulated genes AKA genes induced by rosi (downstream most likely)
umuc9_rosi_upregulated_genes <- umuc9_rosi_siggenes$genes.up[, "Gene Name"]
head(umuc9_rosi_upregulated_genes)

#number of genes upregulated
umuc9_rosi_siggenes$ngenes.up
```

**Define Curated GATA3 Pathway Signature**

```{r}
gata3_sig <- c(
  "ADCY1", "GATA3", "GNAS", "IL13", "IL1A", "IL4", "IL5",
  "JUNB", "MAF", "MAP2K3", "MAP2K6", "MAPK14", "NFATC1",
  "PRKACB", "PRKACG", "PRKAR1A", "PRKAR1B", "PRKAR2A", "PRKAR2B"
)
```

# **Load, Normalize, and Map Mouse RNA-seq Data to Human Gene Symbols**

- Loads parental and MP1 mouse RNA-seq expression matrix
- Removes count, gene ID, and gene biotype columns not needed for scoring
- Keeps one row per mouse gene symbol
- Applies upper-quartile log2 normalization (USES CPM, NOT TPM)
- Converts mouse gene symbols to human orthologs so they can be compared with human PPARG and GATA3 signatures
- Collapses duplicate human ortholog mappings by averaging expression values

```{r}
mouse_cpm_expr <- read.csv(here("data", "parental_mp1_K8HVZP-expression-matrix.csv"), sep=",", header=T)
mouse_cpm_expr <- mouse_cpm_expr[, !grepl("count|^gene_biotype|gene_id", colnames(mouse_cpm_expr)    )    ] #remove count, gene_biotype, gene_id columns
mouse_cpm_expr <- mouse_cpm_expr %>% #remove duplicate reads
  mutate(row_avg = rowMeans(pick(where(is.numeric)), na.rm = TRUE)) %>% #make new column for avg expression across all samples
  arrange(desc(row_avg)) %>%
  distinct(gene_name, .keep_all=TRUE) %>% #keep only read w/ highest avg read across all samples
  dplyr::select(-row_avg) #remove temp column
rownames(mouse_cpm_expr) <- mouse_cpm_expr$gene_name
mouse_cpm_expr$gene_name <- NULL

# Convert cpm to upper quartile log2(cpm+1)
# IMPORTANT: WE USED CPM BUT THEY USED TPM
# SINCE WE ONLY WERE GIVEN CPM AND I DIDN'T WANT TO PROCESS FOR TPM
mouse_uqlog2_expr <- cpm_to_uq_log2(mouse_cpm_expr)

#get mouse symbols from our samples
mouse_gene_symbols <- unique(rownames(mouse_uqlog2_expr))

# get map of mouse:human genes for conversion
mouse_to_human_orthologs <- babelgene::orthologs(
  genes = mouse_gene_symbols,
  species = "mouse",
  human = FALSE,
  top=TRUE
)

#Make new column for human gene ortholog
mouse_uqlog2_expr$human_symbol <- mouse_to_human_orthologs$human_symbol[
  match(rownames(mouse_uqlog2_expr), mouse_to_human_orthologs$symbol)
]

#keep only reads that have human ortholog
sample_cols <- setdiff(colnames(mouse_uqlog2_expr), "human_symbol")

humanized_mouse_expr <- mouse_uqlog2_expr %>%
  filter(!is.na(human_symbol), human_symbol != "") %>% #remove NAs
  group_by(human_symbol) %>%
  summarise(across(all_of(sample_cols), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  column_to_rownames("human_symbol")
```

**Signature Activity Plotting Functions**

- Builds plotting function that can be reused for PPARG, GATA3, or other gene signatures
- Converts signature scores into organized plotting table
- Annotates each sample by model, such as `BBN963` or `UPPL1541`.
- Annotates each sample by subtype (parental or MP1)
- Creates boxplot with jittered sample-level points
- Saves plot to `results/plots/` and also returns plot object

# **Score PPARG Activity in Mouse Samples Using the Public Rosi Signature**

```{r}
# get pparg genes found in rnaseq data
genes_use_pparg <- intersect(umuc9_rosi_upregulated_genes, rownames(humanized_mouse_expr))

#get z score
pparg_gene_z <- t(scale(t(humanized_mouse_expr[genes_use_pparg, , drop = FALSE])))
pparg_score <- colMeans(pparg_gene_z, na.rm = TRUE)

umuc9_pparg_score_df <- format_score(pparg_score, signature_label = "PPARG")

plot_signature_activity(
  score_df = umuc9_pparg_score_df,
  signature_label = "PPARG",
  x_axis_group = "parental",
  stratify_group = "cell_line",
  plot_title="PPARG Activity from UMUC9",
  out_file="pparg_umuc9_activity.png"
)

plot_signature_bbn_vs_uppl_combined(
  umuc9_pparg_score_df,
  signature_label = "PPARG",
  plot_title = "PPARG Activity from UMUC9",
  out_file = "PPARG_UMUC9_BBN_UPPL_comp",
  save_plot = TRUE,
  model_cols = c(
    "BBN" = "#1E3A8A",
    "UPPL" = "#F5A623"
    )
  )

plot_signature_bbn_vs_uppl_pooled(
  score_df=umuc9_pparg_score_df,
  signature_label="PPARG",
  plot_title = "PPARG Signature Activity from UMUC9 in BBN and UPPL Models",
  out_file = "PPARG_UMUC9_signature_activity_BBN_UPPL.png",
  save_plot = TRUE,
  model_cols = c(
    "BBN" = "#1E3A8A",
    "UPPL" = "#F5A623"
  )
)

plot_signature_parental_vs_mp1_combined(
  umuc9_pparg_score_df, 
  signature_label="PPARG",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "PPARG Activity from UMUC9",
  out_file = "PPARG_UMUC9_P_MP1_comp"
)

plot_signature_parental_vs_mp1_pooled(
  umuc9_pparg_score_df,
  signature_label = "PPARG",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "PPARG Signature Activity in UMUC9 Models: Parental vs MP1",
  out_file = "PPARG_signature_activity_UMUC9_parental_vs_MP1"
)
```

**Score GATA3 Pathway Activity in Mouse Samples**

```{r}
genes_use_gata3 <- intersect(gata3_sig, rownames(humanized_mouse_expr))

gata3_gene_z <- t(scale(t(humanized_mouse_expr[genes_use_gata3, , drop = FALSE])))
gata3_score <- colMeans(gata3_gene_z, na.rm = TRUE)


gata3_scores_df <- format_score(gata3_score, signature_label = "GATA3")

plot_signature_activity(
  score_df = gata3_scores_df,
  signature_label = "GATA3",
  x_axis_group = "parental",
  stratify_group = "cell_line",
  plot_title="GATA3 Activity",
  out_file="gata3_activity.png"
)

plot_signature_bbn_vs_uppl(gata3_scores_df,
                           signature_label = "GATA3",
                           parental_filter = "Parental",
                           plot_title = "GATA3 Activity of Parentals",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#0072B2",
                             "UPPL" = "#D55E00"
                           ))

plot_signature_bbn_vs_uppl_combined(gata3_scores_df,
                                    signature_label = "GATA3",
                                    plot_title = "GATA3 Activity by Parental/MP1",
                                    out_file = NULL,
                                    save_plot = TRUE,
                                    model_cols = c(
                                      "BBN" = "#1E3A8A",
                                      "UPPL" = "#F5A623"
                                    ))

plot_signature_bbn_vs_uppl_pooled(
  score_df=gata3_scores_df,
  signature_label="GATA3",
  plot_title = "GATA3 Signature Activity in BBN and UPPL Models",
  out_file = "GATA3_signature_activity_BBN_UPPL.png",
  save_plot = TRUE,
  model_cols = c(
    "BBN" = "#1E3A8A",
    "UPPL" = "#F5A623"
  )
)

plot_signature_parental_vs_mp1_combined(
  gata3_scores_df, 
  signature_label="GATA3",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "GATA3 Signature Activity by Subtype",
  out_file=NULL
)

plot_signature_parental_vs_mp1_pooled(
  gata3_scores_df,
  signature_label = "GATA3",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "GATA3 Signature Activity: Parental vs MP1",
  out_file = "GATA3_signature_activity_parental_vs_MP1"
)
```

# **Identify In-House Rosiglitazone-Responsive Genes**

- u1 = umuc1
- u6 = umuc6
- u14 = umuc14
- rt = rt112

```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)

#load in tpm data and convert to log2(tpm+1)
tpm_umuc_rt_rosi_tpm <- read.delim("~/Documents/GitHub/mouse-models/BASE47/data/rosi_counts_tpm_gene_id.tsv", row.names=1)
tpm_umuc_rt_rosi_tpm <- tpm_umuc_rt_rosi_tpm[, !grepl("_N_", names(tpm_umuc_rt_rosi_tpm))]
tpm_umuc_rt_rosi_log2 <- log2(tpm_umuc_rt_rosi_tpm + 1)

#get ensembl id without . 
umuc_rt_rosi_ensembl_genes <- sub("\\.\\d+$", "", rownames(tpm_umuc_rt_rosi_log2))

#get human gene names mapping
umuc_rt112_gene_symbols <- suppressMessages(
  mapIds(
    org.Hs.eg.db,
    keys = umuc_rt_rosi_ensembl_genes,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
)

# Automatically label:
#   rosi-treatment: 1
#   control: 2
rosi_treatment_labels <- ifelse(grepl("_R_", colnames(tpm_umuc_rt_rosi_log2)), 1,
            ifelse(grepl("_U_", colnames(tpm_umuc_rt_rosi_log2)), 2, NA))

stopifnot(!any(is.na(rosi_treatment_labels)))
stopifnot(length(rosi_treatment_labels) == ncol(tpm_umuc_rt_rosi_log2))

sam_data <- list(
  x = as.matrix(tpm_umuc_rt_rosi_log2),
  y = rosi_treatment_labels,
  geneid = rownames(tpm_umuc_rt_rosi_log2),
  genenames = umuc_rt112_gene_symbols,
  logged2 = TRUE
)

samr_obj <- quiet(
  samr(data=sam_data, resp.type = "Two class unpaired", nperms = 100)
)
delta_table <- quiet(
  samr.compute.delta.table(samr_obj)
  )


# Choose a delta where median FDR < 0.05
# -b/c delta != FDR
delta_df <- as.data.frame(delta_table)
chosen_delta <- delta_df[delta_df[, "median FDR"] < 0.05, "delta"][1]

# get significant genes in rosi vs ctrl
umuc_rt112_siggenes <- samr.compute.siggenes.table(
  samr.obj = samr_obj,
  del = chosen_delta,
  data = sam_data,
  delta.table = delta_table
)
```

# **Score Samples Using In-House Rosi-Responsive Signature**

```{r}
umuc_rt112_pparg_signature <- umuc_rt112_siggenes$genes.up[, "Gene ID"]

umuc_rt112_pparg_genes_in_mouse_expr <- intersect(umuc_rt112_pparg_signature, rownames(humanized_mouse_expr))

#get z score
umuc_rt112_pparg_gene_zscores <- t(scale(t(humanized_mouse_expr[umuc_rt112_pparg_genes_in_mouse_expr, , drop = FALSE])))
umuc_rt112_pparg_scores <- colMeans(umuc_rt112_pparg_gene_zscores, na.rm = TRUE)

umuc_rt112_pparg_scores_df <- format_score(umuc_rt112_pparg_scores, signature_label = "PPARG")

plot_signature_activity(
  score_df = umuc_rt112_pparg_scores_df,
  signature_label = "PPARG",
  x_axis_group = "parental",
  stratify_group = "cell_line",
  plot_title="PPARG Activity",
  out_file="pparg_umuc_rt112_activity.png"
)

plot_signature_bbn_vs_uppl_combined(umuc_rt112_pparg_scores_df,
                                    signature_label = "PPARG",
                                    plot_title = "PPARG Activity from UMUC/RT112",
                                    out_file = "PPARG_UMUC_RT112_BBN_UPPL_comp",
                                    save_plot = TRUE,
                                    model_cols = c(
                                      "BBN" = "#1E3A8A",
                                      "UPPL" = "#F5A623"
                                    ))

plot_signature_bbn_vs_uppl_pooled(
  score_df=umuc_rt112_pparg_scores_df,
  signature_label="PPARG",
  plot_title = "PPARG Signature Activity from UMUC+RT112 in BBN and UPPL Models",
  out_file = "PPARG_UMUC_RT112_signature_activity_BBN_UPPL.png",
  save_plot = TRUE,
  model_cols = c(
    "BBN" = "#1E3A8A",
    "UPPL" = "#F5A623"
  )
)

plot_signature_parental_vs_mp1_combined(
  umuc_rt112_pparg_scores_df, 
  signature_label="PPARG",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "PPARG Activity from UMUC/RT112",
  out_file = "PPARG_UMUC_RT112_P_MP1_comp"
)

plot_signature_parental_vs_mp1_pooled(
  umuc_rt112_pparg_scores_df,
  signature_label = "PPARG",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "PPARG Signature Activity in UMUC and RT112 Models: Parental vs MP1",
  out_file = "PPARG_signature_activity_UMUC_RT112_parental_vs_MP1"
)
```

Run SAM on 4 lines separately:

```{r}
# Cell-line identifiers in the sample names
cell_lines <- c(
  umuc1  = "u1",
  umuc6  = "u6",
  umuc14 = "u4",
  rt112  = "r1"
)

run_cell_line_sam <- function(cell_line_id,
                              expression_data,
                              gene_symbols,
                              fdr_cutoff = 0.05,
                              nperms = 100) {
  
  # Match the cell-line ID as a complete underscore-separated token
  # This prevents "u1" from matching "u14"
  cell_line_pattern <- paste0("(^|_)", cell_line_id, "(_|$)")
  
  keep_samples <- grepl(
    cell_line_pattern,
    colnames(expression_data),
    ignore.case = TRUE
  )
  
  cell_line_expression <- expression_data[, keep_samples, drop = FALSE]
  
  if (ncol(cell_line_expression) == 0) {
    stop("No samples found for cell line: ", cell_line_id)
  }
  
  # Rosi treatment = 1; untreated control = 2
  treatment_labels <- ifelse(
    grepl("_R_", colnames(cell_line_expression)),
    1,
    ifelse(
      grepl("_U_", colnames(cell_line_expression)),
      2,
      NA
    )
  )
  
  if (anyNA(treatment_labels)) {
    stop(
      "Could not assign treatment labels for: ",
      paste(colnames(cell_line_expression)[is.na(treatment_labels)],
            collapse = ", ")
    )
  }
  
  if (!all(c(1, 2) %in% treatment_labels)) {
    stop(
      "Cell line ", cell_line_id,
      " does not contain both rosi and control samples."
    )
  }
  
  sam_data <- list(
    x = as.matrix(cell_line_expression),
    y = treatment_labels,
    geneid = rownames(cell_line_expression),
    genenames = gene_symbols,
    logged2 = TRUE
  )
  
  samr_obj <- quiet(
    samr(
      data = sam_data,
      resp.type = "Two class unpaired",
      nperms = nperms
    )
  )
  
  delta_table <- quiet(
    samr.compute.delta.table(samr_obj)
  )
  
  delta_df <- as.data.frame(delta_table)
  
  eligible_deltas <- delta_df[
    !is.na(delta_df[["median FDR"]]) &
      delta_df[["median FDR"]] < fdr_cutoff,
    "delta"
  ]
  
  if (length(eligible_deltas) == 0) {
    warning(
      "No delta produced a median FDR below ",
      fdr_cutoff,
      " for cell line ",
      cell_line_id
    )
    
    return(list(
      siggenes = NULL,
      chosen_delta = NA_real_,
      delta_table = delta_table,
      samr_obj = samr_obj,
      samples = colnames(cell_line_expression)
    ))
  }
  
  chosen_delta <- eligible_deltas[1]
  
  siggenes <- samr.compute.siggenes.table(
    samr.obj = samr_obj,
    del = chosen_delta,
    data = sam_data,
    delta.table = delta_table
  )
  
  list(
    siggenes = siggenes,
    chosen_delta = chosen_delta,
    delta_table = delta_table,
    samr_obj = samr_obj,
    samples = colnames(cell_line_expression)
  )
}

# Run a separate SAM analysis for each cell line
rosi_sam_results <- lapply(
  cell_lines,
  run_cell_line_sam,
  expression_data = tpm_umuc_rt_rosi_log2,
  gene_symbols = umuc_rt112_gene_symbols,
  fdr_cutoff = 0.05,
  nperms = 100
)

# Four significant-gene table objects
umuc1_siggenes  <- rosi_sam_results$umuc1$siggenes$genes.up[, "Gene Name"]
umuc6_siggenes  <- rosi_sam_results$umuc6$siggenes$genes.up[, "Gene Name"]
umuc14_siggenes <- rosi_sam_results$umuc14$siggenes$genes.up[, "Gene Name"]
rt112_siggenes  <- rosi_sam_results$rt112$siggenes$genes.up[, "Gene Name"]
```

```{r}
# install.packages("ComplexUpset")

library(ComplexUpset)
library(ggplot2)

upregulated_gene_sets <- list(
  UMUC1  = unique(na.omit(umuc1_siggenes)),
  UMUC6  = unique(na.omit(umuc6_siggenes)),
  UMUC14 = unique(na.omit(umuc14_siggenes)),
  RT112  = unique(na.omit(rt112_siggenes)),
  UMUC9 = unique(na.omit(umuc9_rosi_upregulated_genes))
)

all_genes <- sort(unique(unlist(upregulated_gene_sets)))

upset_data <- data.frame(
  gene = all_genes,
  UMUC1 = all_genes %in% upregulated_gene_sets$UMUC1,
  UMUC6 = all_genes %in% upregulated_gene_sets$UMUC6,
  UMUC14 = all_genes %in% upregulated_gene_sets$UMUC14,
  RT112 = all_genes %in% upregulated_gene_sets$RT112,
  UMUC9 = all_genes %in% upregulated_gene_sets$UMUC9
)

upset_plot <- ComplexUpset::upset(
  upset_data,
  intersect = c("UMUC1", "UMUC6", "UMUC14", "RT112", "UMUC9"),
  name = "Cell line",
  min_size = 0,
  sort_intersections_by = c("degree", "cardinality"),
  base_annotations = list(
    "Intersection size" =
      intersection_size(counts = TRUE)
  ),
  set_sizes = (
    upset_set_size() +
      ylab("Upregulated genes")
  )
) +
  ggtitle("Overlap of significantly upregulated genes")

upset_plot

ggsave(
  here("results","plots","rosi_upregulated_genes_upset_plot.png"),
  upset_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

Overlapping Significantly Upregulated Genes between UMUC9 and UMUCs+RT112

```{r}
library(ggforce)
library(ggrepel)


umuc_rt112_pparg_signature

genes_use_pparg
highlight_genes <- c("TP53", "BRCA1", "NECTIN4", "FABP4", "GATA3", "KRT6", "KRT14", "KRT20", "PPARG")

umuc9_only <- setdiff(genes_use_pparg, umuc_rt112_pparg_signature)
umuc_rt112_only <- setdiff(umuc_rt112_pparg_signature, genes_use_pparg)
both_umuc9_umuc_rt112   <- intersect(genes_use_pparg, umuc_rt112_only)

counts <- data.frame(
  region = c("UMUC9 only", "Both", "UMUC+RT112 only"),
  x = c(-0.8, 0, 0.8),
  y = c(0, 0, 0),
  label = c(length(umuc9_only), length(both_umuc9_umuc_rt112), length(umuc_rt112_only))
)

highlight_df <- data.frame(gene = highlight_genes) %>%
  filter(gene %in% union(umuc9_only, umuc_rt112_only)) %>%
  mutate(
    region = case_when(
      gene %in% both_umuc9_umuc_rt112 ~ "Both",
      gene %in% umuc9_only ~ "A only",
      gene %in% umuc_rt112_only ~ "B only"
    ),
    x = case_when(
      region == "A only" ~ -0.8,
      region == "Both" ~ 0,
      region == "B only" ~ 0.8
    ),
    y = row_number() * 0.15
  )


ggplot() +
  geom_circle(
    aes(x0 = -0.5, y0 = 0, r = 1),
    fill = "skyblue",
    alpha = 0.35,
    color = "black"
  ) +
  geom_circle(
    aes(x0 = 0.5, y0 = 0, r = 1),
    fill = "salmon",
    alpha = 0.35,
    color = "black"
  ) +
  geom_text(
    data = counts,
    aes(x = x, y = y, label = label),
    size = 7
  ) +
  geom_text_repel(
    data = highlight_df,
    aes(x = x, y = y, label = gene),
    size = 4,
    fontface = "bold",
    box.padding = 0.4
  ) +
  annotate("text", x = -0.9, y = 1.15, label = "UMUC9", size = 5) +
  annotate("text", x = 0.9, y = 1.15, label = "UMUC+RT112", size = 5) +
  coord_fixed() +
  theme_void()
```

Check if genes upregulated in OUR rosi data (quickly)

```{r}
library(readr)
library(dplyr)
library(purrr)

check_upregulation <- function(row, pseudocount = 0) {
  cell_lines <- c("R1", "U1", "U4", "U6")
  
  map_dfr(cell_lines, function(cell) {
    rosi_cols <- grep(paste0("^", cell, "_R_"), names(row), value = TRUE)
    ctrl_cols <- grep(paste0("^", cell, "_U_"), names(row), value = TRUE)
    
    rosi_mean <- mean(as.numeric(unlist(row[rosi_cols])), na.rm = TRUE)
    ctrl_mean <- mean(as.numeric(unlist(row[ctrl_cols])), na.rm = TRUE)
    
    fold_change <- (rosi_mean + pseudocount) /
      (ctrl_mean + pseudocount)
    
    tibble(
      feature_id  = as.character(row[["feature_id"]][1]),
      cell_line   = cell,
      rosi_mean   = rosi_mean,
      control_mean = ctrl_mean,
      fold_change = fold_change,
      log2FC      = log2(fold_change),
      upregulated = rosi_mean > ctrl_mean
    )
  })
}

counts <- read_tsv(
  "/Users/timothy/Documents/GitHub/mouse-models/data/ROSI_UMUC9_UMUC14_counts_tpm_gene_id.tsv"
)


N4 <- counts %>%
  filter(feature_id == "ENSG00000170323.8") %>% #NECTIN4
  check_upregulation(pseudocount = 1)

PPARG <- counts %>%
  filter(feature_id == "ENSG00000132170.20") %>% #NECTIN4
  check_upregulation(pseudocount = 1)

FABP4 <- counts %>%
  filter(feature_id == "ENSG00000170323.8") %>% #NECTIN4
  check_upregulation(pseudocount = 1)
```
