---
title: "GATA3/PPARG Scoring"
format: html
editor: visual
---

Folder structure: GATA3_PPARG_Scoring/ \|-- data/ \|-- results/ \| `-- plots/`-- scripts/ \`-- run_gata3_pparg_files/ ---

# Setup

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(here)
# here::i_am("README.Rmd")

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("GEOquery", "impute"))
library(samr)
library(GEOquery)
library(limma)
library(tidyverse)

# =========
# Functions
# =========

#Convert to 90th percentile of log(tpm+1)
cpm_to_uq_log2 <- function(cpm_matrix) {
  uq <- apply(cpm_matrix, 2, function(x) quantile(x[x > 0], 0.75))
  scaled <- sweep(cpm_matrix, 2, uq, "/") * 1000
  log2(scaled + 1)
}

# Suppress permutations output
quiet <- function(expr) {
  suppressMessages(suppressWarnings({
    capture.output(result <- eval.parent(substitute(expr)))
    result
  }))
}
```

# Derive Public Rosiglitazone-Induced PPARγ Pathway Signature from GSE47993

1.  Downloads public FABP4-modulation GEO dataset GSE47993
2.  Subset to only UMUC9 control and rosi-treated samples
    1.  Uses rosiglitazone treatment as a PPARγ activation condition
3.  Map Illumina Probe IDs to gene symbols
4.  Collapse multiple probes mapping to the same gene by averaging expression values
5.  Uses SAM to identify genes significantly upregulated after rosi treatment

```{r}
gse47993_dataset <- getGEO("GSE47993", GSEMatrix = TRUE)[[1]]
gse47993_expr <- exprs(gse47993_dataset)

umuc9_rosi_expr <- gse47993_expr[, c("GSM1164283","GSM1164284","GSM1164285",  # control samples
                       "GSM1164286","GSM1164287","GSM1164288") # rosi treated samples
]

# Get gene names from illumina probe IDs (they used IlluminaHT-12 V4.0 expression beadchip)
gse47993_annotation <- fData(gse47993_dataset)
probe_gene_symbols <- gse47993_annotation[rownames(umuc9_rosi_expr), "Symbol"]

# keep probes with valid gene symbols
umuc9_expr_with_symbols <- umuc9_rosi_expr[!is.na(probe_gene_symbols) & probe_gene_symbols != "", ]
probe_gene_symbols <- probe_gene_symbols[!is.na(probe_gene_symbols) & probe_gene_symbols != ""]
# Collapse multiple probes per gene
umuc9_gene_expr <- avereps(umuc9_expr_with_symbols, ID = probe_gene_symbols) # now have gene expression x sample matrix
rosi_group_labels <- c(1,1,1,2,2,2) #1=ctrl, #2=rosi-treated

umuc9_sam_input <- list(
  x = umuc9_gene_expr, 
  y = rosi_group_labels,
  geneid = rownames(umuc9_gene_expr),
  genenames = rownames(umuc9_gene_expr),
  logged2 = TRUE
)

umuc9_sam_fit <- quiet(
  samr(
    data=umuc9_sam_input, resp.type = "Two class unpaired", nperms = 100)
)
umuc9_delta_table <- quiet(
  samr.compute.delta.table(umuc9_sam_fit)
)

# Choose a delta where median FDR < 0.05
# -b/c delta != FDR
sam_delta_results_umuc9 <- as.data.frame(umuc9_delta_table)
fdr05_delta_umuc9 <- sam_delta_results_umuc9[sam_delta_results_umuc9[, "median FDR"] < 0.05, "delta"][1]

# get significant genes in rosi vs ctrl
umuc9_rosi_siggenes <- samr.compute.siggenes.table(
  samr.obj = umuc9_sam_fit,
  del = fdr05_delta_umuc9,
  data = umuc9_sam_input,
  delta.table = umuc9_delta_table
)
```

**Extract Rosi-Induced PPARG Target Genes**

- Saves significantly upregulated rosiglitazone-responsive genes as PPARG activity signature
- These genes represent downstream transcriptional changes associated with PPARG activation

```{r}
# Significant upregulated genes AKA genes induced by rosi (downstream most likely)
umuc9_rosi_upregulated_genes <- umuc9_rosi_siggenes$genes.up[, "Gene Name"]
head(umuc9_rosi_upregulated_genes)

#number of genes upregulated
umuc9_rosi_siggenes$ngenes.up
```

**Define Curated GATA3 Pathway Signature**

```{r}
gata3_sig <- c(
  "ADCY1", "GATA3", "GNAS", "IL13", "IL1A", "IL4", "IL5",
  "JUNB", "MAF", "MAP2K3", "MAP2K6", "MAPK14", "NFATC1",
  "PRKACB", "PRKACG", "PRKAR1A", "PRKAR1B", "PRKAR2A", "PRKAR2B"
)
```

# **Load, Normalize, and Map Mouse RNA-seq Data to Human Gene Symbols**

- Loads parental and MP1 mouse RNA-seq expression matrix
- Removes count, gene ID, and gene biotype columns not needed for scoring
- Keeps one row per mouse gene symbol
- Applies upper-quartile log2 normalization (USES CPM, NOT TPM)
- Converts mouse gene symbols to human orthologs so they can be compared with human PPARG and GATA3 signatures
- Collapses duplicate human ortholog mappings by averaging expression values

```{r}
mouse_cpm_expr <- read.csv(here("data", "parental_mp1_K8HVZP-expression-matrix.csv"), sep=",", header=T)
mouse_cpm_expr <- mouse_cpm_expr[, !grepl("count|^gene_biotype|gene_id", colnames(mouse_cpm_expr)    )    ] #remove count, gene_biotype, gene_id columns
mouse_cpm_expr <- mouse_cpm_expr %>% #remove duplicate reads
  mutate(row_avg = rowMeans(pick(where(is.numeric)), na.rm = TRUE)) %>% #make new column for avg expression across all samples
  arrange(desc(row_avg)) %>%
  distinct(gene_name, .keep_all=TRUE) %>% #keep only read w/ highest avg read across all samples
  dplyr::select(-row_avg) #remove temp column
rownames(mouse_cpm_expr) <- mouse_cpm_expr$gene_name
mouse_cpm_expr$gene_name <- NULL

# Convert cpm to upper quartile log2(cpm+1)
# IMPORTANT: WE USED CPM BUT THEY USED TPM
# SINCE WE ONLY WERE GIVEN CPM AND I DIDN'T WANT TO PROCESS FOR TPM
mouse_uqlog2_expr <- cpm_to_uq_log2(mouse_cpm_expr)

#get mouse symbols from our samples
mouse_gene_symbols <- unique(rownames(mouse_uqlog2_expr))

# get map of mouse:human genes for conversion
mouse_to_human_orthologs <- babelgene::orthologs(
  genes = mouse_gene_symbols,
  species = "mouse",
  human = FALSE,
  top=TRUE
)

#Make new column for human gene ortholog
mouse_uqlog2_expr$human_symbol <- mouse_to_human_orthologs$human_symbol[
  match(rownames(mouse_uqlog2_expr), mouse_to_human_orthologs$symbol)
]

#keep only reads that have human ortholog
sample_cols <- setdiff(colnames(mouse_uqlog2_expr), "human_symbol")

humanized_mouse_expr <- mouse_uqlog2_expr %>%
  filter(!is.na(human_symbol), human_symbol != "") %>% #remove NAs
  group_by(human_symbol) %>%
  summarise(across(all_of(sample_cols), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  column_to_rownames("human_symbol")
```

**Signature Activity Plotting Functions**

- Builds plotting function that can be reused for PPARG, GATA3, or other gene signatures
- Converts signature scores into organized plotting table
- Annotates each sample by model, such as `BBN963` or `UPPL1541`.
- Annotates each sample by subtype (parental or MP1)
- Creates boxplot with jittered sample-level points
- Saves plot to `results/plots/` and also returns plot object

```{r}
# Formats z-score into dataframe that can be used for plotting

format_score <- function(score_vec, signature_label) {
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single string.")
  }
  
  if (is.null(names(score_vec))) {
    stop("score_vec must be a named vector.")
  }
  
  score_col <- paste0(signature_label, "_z_score")
  sample_names <- names(score_vec)
  
  score_df <- data.frame(
    cell_line = sub("_.*", "", sample_names),
    parental = ifelse(grepl("_PARENTAL_", sample_names), "Parental", "MP1"),
    score_vec,
    check.names = FALSE
  )
  
  names(score_df)[names(score_df) == "score_vec"] <- score_col
  
  return(score_df)
}

# Create 2 complementary plots of signature activity
# 1. grouped_dot_plot
#    Shows Parental and MP1 samples separately for each cell line
# 2. interaction_plot
#    Connects the group means for each cell line, making it easier
#    to see whether the Parental-to-MP1 change differs by cell line

plot_signature_activity <- function(
    score_df,
    signature_label,
    x_axis_group = "parental",
    stratify_group = "cell_line",
    plot_title = NULL,
    y_axis_label = NULL,
    out_file = NULL,
    save_plot = TRUE
) {
  
  # ---------------------------------------------------------------------------
  # Validate the signature label
  # ---------------------------------------------------------------------------
  
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single non-empty string.")
  }
  
  # Construct the expected score-column name.
  #
  # Example:
  #   signature_label = "PPARG"
  #   score_col = "PPARG_z_score"
  score_col <- paste0(signature_label, "_z_score")
  
  if (!score_col %in% colnames(score_df)) {
    stop(
      paste0(
        "Column '",
        score_col,
        "' not found in score_df."
      )
    )
  }
  
  # ---------------------------------------------------------------------------
  # Validate the grouping variables
  # ---------------------------------------------------------------------------
  
  # These are the only two columns that can control the plot grouping.
  valid_groups <- c("cell_line", "parental")
  
  if (!x_axis_group %in% valid_groups) {
    stop(
      paste0(
        "x_axis_group must be either 'cell_line' or 'parental'. ",
        "Received: '",
        x_axis_group,
        "'."
      )
    )
  }
  
  if (!stratify_group %in% valid_groups) {
    stop(
      paste0(
        "stratify_group must be either 'cell_line' or 'parental'. ",
        "Received: '",
        stratify_group,
        "'."
      )
    )
  }
  
  # The same column cannot control both plot roles.
  #
  # Valid:
  #   x_axis_group = "parental"
  #   stratify_group = "cell_line"
  #
  # Also valid:
  #   x_axis_group = "cell_line"
  #   stratify_group = "parental"
  #
  # Invalid:
  #   x_axis_group = "parental"
  #   stratify_group = "parental"
  if (x_axis_group == stratify_group) {
    stop(
      paste0(
        "x_axis_group and stratify_group must be different. ",
        "Use one of the following combinations:\n",
        "  x_axis_group = 'parental', stratify_group = 'cell_line'\n",
        "  x_axis_group = 'cell_line', stratify_group = 'parental'"
      )
    )
  }
  
  # Confirm that both selected columns exist in the data frame.
  if (!x_axis_group %in% colnames(score_df)) {
    stop(
      paste0(
        "Column '",
        x_axis_group,
        "' not found in score_df."
      )
    )
  }
  
  if (!stratify_group %in% colnames(score_df)) {
    stop(
      paste0(
        "Column '",
        stratify_group,
        "' not found in score_df."
      )
    )
  }
  
  # ---------------------------------------------------------------------------
  # Create default labels
  # ---------------------------------------------------------------------------
  
  if (is.null(plot_title)) {
    plot_title <- paste(signature_label, "Activity")
  }
  
  if (is.null(y_axis_label)) {
    y_axis_label <- paste(
      signature_label,
      "Signature Z-Score"
    )
  }
  
  # Human-readable labels for plot legends.
  group_labels <- c(
    "cell_line" = "Cell line",
    "parental" = "Condition"
  )
  
  # ---------------------------------------------------------------------------
  # Define category order
  # ---------------------------------------------------------------------------
  
  # These levels determine the order displayed on the plot.
  group_levels <- list(
    parental = c(
      "Parental",
      "MP1"
    ),
    cell_line = c(
      "BBN963",
      "UPPL1541"
    )
  )
  
  # ---------------------------------------------------------------------------
  # Define colors
  # ---------------------------------------------------------------------------
  
  # Colors for each possible grouping variable.
  #
  # The function automatically selects the correct palette depending
  # on which column is used for each plot role.
  group_colors <- list(
    
    parental = c(
      "Parental" = "#18AEB5",
      "MP1" = "#F05A50"
    ),
    
    cell_line = c(
      "BBN963" = "#0072B2",
      "UPPL1541" = "#D55E00"
    )
  )
  
  # Colors for the variable displayed on the x-axis.
  x_axis_colors <- group_colors[[x_axis_group]]
  
  # Colors for the variable represented by interaction lines.
  series_colors <- group_colors[[stratify_group]]
  
  # ---------------------------------------------------------------------------
  # Prepare the data
  # ---------------------------------------------------------------------------
  
  score_df <- score_df %>%
    mutate(
      
      # Convert the x-axis column to a factor and set its category order.
      "{x_axis_group}" := factor(
        .data[[x_axis_group]],
        levels = group_levels[[x_axis_group]]
      ),
      
      # Convert the series column to a factor and set its category order.
      "{stratify_group}" := factor(
        .data[[stratify_group]],
        levels = group_levels[[stratify_group]]
      )
    ) %>%
    
    # Remove rows missing information required for plotting.
    filter(
      !is.na(.data[[x_axis_group]]),
      !is.na(.data[[stratify_group]]),
      !is.na(.data[[score_col]])
    )
  
  # Stop if filtering removed all rows.
  if (nrow(score_df) == 0) {
    stop(
      "No complete rows remain after filtering the grouping and score columns."
    )
  }
  
  # ---------------------------------------------------------------------------
  # Create grouped dot plot
  # ---------------------------------------------------------------------------
  
  # The grouped dot plot:
  #
  #   - displays x_axis_group on the x-axis
  #   - colors points by x_axis_group
  #   - creates one panel for each stratify_group value
  #
  # Default orientation:
  #   x-axis: Parental and MP1
  #   panels: BBN963 and UPPL1541
  #
  # Swapped orientation:
  #   x-axis: BBN963 and UPPL1541
  #   panels: Parental and MP1
  grouped_dot_plot <- ggplot(
    score_df,
    aes(
      x = .data[[x_axis_group]],
      y = .data[[score_col]],
      color = .data[[x_axis_group]]
    )
  ) +
    
    # Add a reference line at a z-score of zero.
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    
    # Plot individual sample scores.
    geom_jitter(
      width = 0.08,
      height = 0,
      size = 2.8,
      alpha = 0.9
    ) +
    
    # Display the group mean as a thick horizontal marker.
    stat_summary(
      fun = mean,
      geom = "point",
      shape = 95,
      size = 10,
      linewidth = 1.2
    ) +
    
    # Create one panel for each series-group value.
    facet_wrap(
      vars(.data[[stratify_group]]),
      nrow = 1
    ) +
    
    # Apply the palette corresponding to the x-axis variable.
    scale_color_manual(
      values = x_axis_colors,
      drop = FALSE
    ) +
    
    labs(
      title = plot_title,
      x = NULL,
      y = y_axis_label,
      color = group_labels[[x_axis_group]]
    ) +
    
    theme_classic(base_size = 12) +
    
    theme(
      plot.title = element_text(face = "bold"),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )
  
  # ---------------------------------------------------------------------------
  # Calculate group means
  # ---------------------------------------------------------------------------
  
  # Calculate the mean for every combination of the two variables.
  #
  # These combinations are the same regardless of which variable is
  # displayed on the x-axis.
  mean_df <- score_df %>%
    group_by(
      .data[[stratify_group]],
      .data[[x_axis_group]]
    ) %>%
    summarise(
      mean_score = mean(
        .data[[score_col]],
        na.rm = TRUE
      ),
      .groups = "drop"
    )
  
  # ---------------------------------------------------------------------------
  # Create interaction plot
  # ---------------------------------------------------------------------------
  
  # The interaction plot:
  #
  #   - displays x_axis_group on the x-axis
  #   - creates one line for each stratify_group value
  #   - colors points and lines by stratify_group
  #
  # Default orientation:
  #   x-axis: Parental and MP1
  #   lines: BBN963 and UPPL1541
  #
  # Swapped orientation:
  #   x-axis: BBN963 and UPPL1541
  #   lines: Parental and MP1
  interaction_plot <- ggplot() +
    
    # Add a reference line at a z-score of zero.
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    
    # Display individual observations faintly in the background.
    geom_jitter(
      data = score_df,
      aes(
        x = .data[[x_axis_group]],
        y = .data[[score_col]],
        color = .data[[stratify_group]]
      ),
      width = 0.08,
      height = 0,
      size = 2.4,
      alpha = 0.45
    ) +
    
    # Connect the group means.
    geom_line(
      data = mean_df,
      aes(
        x = .data[[x_axis_group]],
        y = mean_score,
        group = .data[[stratify_group]],
        color = .data[[stratify_group]]
      ),
      linewidth = 1.1
    ) +
    
    # Add a point at each group mean.
    geom_point(
      data = mean_df,
      aes(
        x = .data[[x_axis_group]],
        y = mean_score,
        color = .data[[stratify_group]]
      ),
      size = 3.5
    ) +
    
    # Apply the palette corresponding to the series variable.
    scale_color_manual(
      values = series_colors,
      drop = FALSE
    ) +
    
    labs(
      title = paste(
        plot_title,
        "\nInteraction View"
      ),
      x = NULL,
      y = y_axis_label,
      color = group_labels[[stratify_group]]
    ) +
    
    theme_classic(base_size = 12) +
    
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "right"
    )
  
  # ---------------------------------------------------------------------------
  # Save plots
  # ---------------------------------------------------------------------------
  
  if (save_plot) {
    
    if (is.null(out_file)) {
      
      # Convert the signature name into a filename-safe value.
      safe_label <- signature_label %>%
        str_replace_all("[^A-Za-z0-9]+", "_") %>%
        str_replace_all("^_|_$", "")
      
      # Include the orientation in the filename.
      #
      # This prevents the swapped version from overwriting plots
      # created using the default orientation.
      out_prefix <- paste0(
        safe_label,
        "_x_",
        x_axis_group,
        "_by_",
        stratify_group
      )
      
    } else {
      
      # Remove an extension supplied by the user.
      out_prefix <- tools::file_path_sans_ext(out_file)
    }
    
    out_dir <- here::here(
      "results",
      "plots"
    )
    
    dir.create(
      out_dir,
      recursive = TRUE,
      showWarnings = FALSE
    )
    
    ggsave(
      filename = file.path(
        out_dir,
        paste0(
          out_prefix,
          "_grouped_dot_plot.png"
        )
      ),
      plot = grouped_dot_plot,
      width = 6,
      height = 4,
      dpi = 300
    )
    
    ggsave(
      filename = file.path(
        out_dir,
        paste0(
          out_prefix,
          "_interaction_plot.png"
        )
      ),
      plot = interaction_plot,
      width = 5,
      height = 4,
      dpi = 300
    )
  }
  
  # Return the plots and calculated means.
  return(
    list(
      grouped_dot_plot = grouped_dot_plot,
      interaction_plot = interaction_plot,
      means = mean_df
    )
  )
}

# Compare BBN and UPPL models after selecting ONE condition,
# such as only Parental samples or only MP1 samples
# Ex. it'll show the signature scores for GATA3 between BBN and UPPL models for ONLY Parental samples

#ANSWERS: Among Parental samples, is signature activity different between the BBN and UPPL model families?

plot_signature_bbn_vs_uppl <- function(
    score_df, #df that contains signature score
    signature_label, #PPARG/GATA3
    parental_filter, #"Parental" or "MP1"
    plot_title = NULL, # optional custom title for plot
    out_file = NULL, #optional output file name
    save_plot = TRUE,
    model_cols = c( # assign colors for plots
      "BBN" = "#0072B2",
      "UPPL" = "#D55E00"
      ))
  {
  
  # confirm signature label is valid
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single string.")
  }
  
  score_col <- paste0(signature_label, "_z_score")
  if (!score_col %in% colnames(score_df)) {
    stop(paste0("Column '", score_col, "' not found in score_df."))
  }
  
  if (!"cell_line" %in% colnames(score_df)) {
    stop("Column 'cell_line' not found in score_df.")
  }
  
  if (!"parental" %in% colnames(score_df)) {
    stop("Column 'parental' not found in score_df.")
  }
  
  if (is.null(plot_title)) {
    plot_title <- paste0(signature_label, " Activity: ", parental_filter, " Only")
  }

  # prepare data for plotting
  score_df <- score_df %>%
    #keep only selected condition
    filter(parental == parental_filter) %>% #ex. keep only parental samples
    mutate(
      # group cell lines into BBN or UPPL
      cell_model = case_when(
        grepl("^BBN", cell_line) ~ "BBN",
        grepl("^UPPL", cell_line) ~ "UPPL",
        TRUE ~ NA_character_
      ),
      cell_model = factor(cell_model, levels = c("BBN", "UPPL"))
    ) %>%
    filter(
      !is.na(cell_model),
      !is.na(.data[[score_col]])
    )
  
  # create BBN vs UPPL activity plot
  activity_plot <- ggplot(
    score_df,
    aes(
      x = cell_model,
      y = .data[[score_col]],
      color = cell_model
    )
  ) +
    # add horizontal reference line @z-score=0
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    #plot individual sample z scores
    geom_jitter(
      width = 0.08,
      height = 0,
      size = 2.8,
      alpha = 0.9
    ) +
    # add mean score as a thick horizontal marker
    stat_summary(
      fun = mean,
      geom = "point",
      shape = 95,
      size = 10,
      linewidth = 1.2
    ) +
    # apply colors for BBN and UPPL
    scale_color_manual(values = model_cols, drop = FALSE) +
    labs(
      title = plot_title,
      x = NULL,
      y = paste(signature_label, "Signature Z-Score"),
      color = NULL
    ) +
    # plot title, axis labels, legend title
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "right"
    )
  # calculate mean signature score for BBN
  mean_df <- score_df %>%
    group_by(cell_model) %>%
    summarise(
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      .groups = "drop"
    )
  # save plot when save_plot=TRUE
  if (save_plot) {
    
    if (is.null(out_file)) {
      safe_label <- signature_label %>%
        str_replace_all("[^A-Za-z0-9]+", "_") %>%
        str_replace_all("^_|_$", "")
      
      safe_condition <- parental_filter %>%
        str_replace_all("[^A-Za-z0-9]+", "_") %>%
        str_replace_all("^_|_$", "")
      
      out_prefix <- paste0(safe_label, "_", safe_condition, "_bbn_vs_uppl")
    } else {
      out_prefix <- tools::file_path_sans_ext(out_file)
    }
    
    out_dir <- here::here("results", "plots")
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    
    ggsave(
      filename = file.path(out_dir, paste0(out_prefix, ".png")),
      plot = activity_plot,
      width = 4.5,
      height = 4,
      dpi = 400
    )
  }
  # return both plot and calculated group means
  return(list(
    activity_plot = activity_plot,
    means = mean_df
  ))
}

# Compare Parental and MP1 models after selecting ONE condition,
# such as only BBN samples or only UPPL samples
# Ex. it'll show the signature scores for GATA3 between Parental and MP1 models for ONLY BBN samples

#ANSWERS: Within BBN subtype, is signature activity different between Parental and MP1 samples?

plot_signature_parental_vs_mp1 <- function(score_df, # input df from z-scores
                                           signature_label, #gata3/pparg
                                           cell_model_filter = "BBN", #subtype: BBN/UPPL 
                                           x_group = "parental",
                                           color_group = "parental",
                                           plot_title = NULL, #custom plot title
                                           out_file = NULL, # saves the plots
                                           save_plot = TRUE,
                                           condition_cols = c(
                                             "Parental" = "#18AEB5", #color for parental
                                             "MP1" = "#F05A50" #color for MP1
                                           )) {
  
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single string.")
  }
  
  score_col <- paste0(signature_label, "_z_score")
  
  if (!score_col %in% colnames(score_df)) {
    stop(paste0("Column '", score_col, "' not found in score_df."))
  }
  
  if (!"cell_line" %in% colnames(score_df)) {
    stop("Column 'cell_line' not found in score_df.")
  }
  
  if (!"parental" %in% colnames(score_df)) {
    stop("Column 'parental' not found in score_df.")
  }
  
  if (is.null(plot_title)) {
    plot_title <- paste0(signature_label, " Activity: ", cell_model_filter, " Only")
  }
    y_axis_label <- paste(signature_label, "Signature Z-Score")
  
  score_df <- score_df %>%
    mutate(
      cell_model = case_when(
        grepl("^BBN", cell_line) ~ "BBN",
        grepl("^UPPL", cell_line) ~ "UPPL",
        TRUE ~ NA_character_
      ),
      parental = factor(parental, levels = c("Parental", "MP1"))
    ) %>%
    filter(
      cell_model == cell_model_filter,
      !is.na(parental),
      !is.na(.data[[score_col]])
    )
  
  activity_plot <- ggplot(
    score_df,
    aes(
      x = .data[[x_group]],
      y = .data[[score_col]],
      color = .data[[color_group]]
    )
  ) +
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    geom_jitter(
      width = 0.08,
      height = 0,
      size = 2.8,
      alpha = 0.9
    ) +
    stat_summary(
      fun = mean,
      geom = "point",
      shape = 95,
      size = 10,
      linewidth = 1.2
    ) +
    scale_color_manual(values = condition_cols, drop = FALSE) +
    labs(
      title = plot_title,
      x = NULL,
      y = y_axis_label,
      color = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "right"
    )
  
  mean_df <- score_df %>%
    group_by(.data[[x_group]]) %>%
    summarise(
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  if (save_plot) {
    
    if (is.null(out_file)) {
      safe_label <- signature_label %>%
        str_replace_all("[^A-Za-z0-9]+", "_") %>%
        str_replace_all("^_|_$", "")
      
      out_prefix <- paste0(safe_label, "_", cell_model_filter, "_parental_vs_mp1")
    } else {
      out_prefix <- tools::file_path_sans_ext(out_file)
    }
    
    out_dir <- here::here("results", "plots")
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    
    ggsave(
      filename = file.path(out_dir, paste0(out_prefix, ".png")),
      plot = activity_plot,
      width = 4.5,
      height = 4,
      dpi = 300
    )
  }
  
  return(list(
    activity_plot = activity_plot,
    means = mean_df
  ))
}

plot_signature_parental_vs_mp1_combined <- function(
    score_df,                          # Data frame containing metadata and z-scores
    signature_label,                   # Signature name, such as "GATA3" or "PPARG"
    cell_models = c("BBN", "UPPL"),     # Models shown as separate subplots
    plot_title = NULL,                 # Optional custom plot title
    out_file = NULL,                   # Optional output filename
    save_plot = TRUE,                  # Whether to save the plot
    condition_cols = c(
      "Parental" = "#18AEB5",          # Color for Parental samples
      "MP1" = "#F05A50"                # Color for MP1 samples
    )
) {
  
  # Check that signature_label is one non-empty string.
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single non-empty string.")
  }
  
  # Construct the expected score-column name.
  #
  # Example:
  #   signature_label = "PPARG"
  #   score_col = "PPARG_z_score"
  score_col <- paste0(signature_label, "_z_score")
  
  # Check that all required columns are present.
  required_cols <- c(
    "cell_line",
    "parental",
    score_col
  )
  
  missing_cols <- setdiff(
    required_cols,
    colnames(score_df)
  )
  
  if (length(missing_cols) > 0) {
    stop(
      paste0(
        "The following required columns are missing: ",
        paste(missing_cols, collapse = ", ")
      )
    )
  }
  
  # Check that cell_models contains only supported model names.
  valid_models <- c("BBN", "UPPL")
  
  if (!all(cell_models %in% valid_models)) {
    stop(
      paste0(
        "cell_models can only contain: ",
        paste(valid_models, collapse = ", ")
      )
    )
  }
  
  # Remove duplicated model names while preserving their order.
  cell_models <- unique(cell_models)
  
  # Create a default plot title.
  if (is.null(plot_title)) {
    plot_title <- paste(
      signature_label,
      "Activity: Parental vs MP1"
    )
  }
  
  # Create the y-axis label.
  y_axis_label <- paste(
    signature_label,
    "Signature Z-Score"
  )
  
  # Prepare the data for plotting.
  plot_df <- score_df %>%
    dplyr::mutate(
      
      # Assign each cell line to the BBN or UPPL model family.
      cell_model = dplyr::case_when(
        grepl("^BBN", cell_line) ~ "BBN",
        grepl("^UPPL", cell_line) ~ "UPPL",
        TRUE ~ NA_character_
      ),
      
      # Display Parental before MP1.
      parental = factor(
        parental,
        levels = c("Parental", "MP1")
      ),
      
      # Control the order of the subplot panels.
      cell_model = factor(
        cell_model,
        levels = cell_models
      )
    ) %>%
    
    # Keep only the requested models and complete observations.
    dplyr::filter(
      cell_model %in% cell_models,
      !is.na(cell_model),
      !is.na(parental),
      !is.na(.data[[score_col]])
    )
  
  # Stop if no valid observations remain.
  if (nrow(plot_df) == 0) {
    stop("No valid observations remain after filtering.")
  }
  
  # Create one subplot for each cell model.
  activity_plot <- ggplot2::ggplot(
    plot_df,
    ggplot2::aes(
      x = parental,
      y = .data[[score_col]],
      color = parental
    )
  ) +
    
    # Add a reference line at a z-score of zero.
    ggplot2::geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    
    # Plot individual sample scores.
    ggplot2::geom_jitter(
      width = 0.08,
      height = 0,
      size = 2.8,
      alpha = 0.9
    ) +
    
    # Add the mean as a thick horizontal marker.
    ggplot2::stat_summary(
      fun = mean,
      geom = "point",
      shape = 95,
      size = 10,
      linewidth = 1.2
    ) +
    
    # Create separate BBN and UPPL subplots.
    ggplot2::facet_wrap(
      ggplot2::vars(cell_model),
      nrow = 1
    ) +
    
    # Apply the Parental and MP1 colors.
    ggplot2::scale_color_manual(
      values = condition_cols,
      drop = FALSE
    ) +
    
    # Add plot labels.
    ggplot2::labs(
      title = plot_title,
      x = NULL,
      y = y_axis_label,
      color = NULL
    ) +
    
    ggplot2::theme_classic(base_size = 12) +
    
    ggplot2::theme(
      plot.title = ggplot2::element_text(
        face = "bold"
      ),
      
      # Add more horizontal space between subplots.
      panel.spacing.x = grid::unit(
        1.5,
        "lines"
      ),
      
      # Draw a border around each subplot.
      panel.border = ggplot2::element_rect(
        color = "grey60",
        fill = NA,
        linewidth = 0.7
      ),
      
      # Add a visible background and border to each subplot title.
      strip.background = ggplot2::element_rect(
        fill = "grey95",
        color = "grey60",
        linewidth = 0.7
      ),
      
      # Format the BBN and UPPL subplot labels.
      strip.text = ggplot2::element_text(
        face = "bold",
        size = 12,
        margin = ggplot2::margin(
          t = 5,
          r = 5,
          b = 5,
          l = 5
        )
      ),
      
      legend.position = "right"
    )
  
  # Calculate the mean score for each model and condition.
  mean_df <- plot_df %>%
    dplyr::group_by(
      cell_model,
      parental
    ) %>%
    dplyr::summarise(
      mean_score = mean(
        .data[[score_col]],
        na.rm = TRUE
      ),
      .groups = "drop"
    )
  
  # Save the combined faceted plot.
  if (save_plot) {
    
    if (is.null(out_file)) {
      
      # Make the signature label safe for use in a filename.
      safe_label <- signature_label %>%
        stringr::str_replace_all(
          "[^A-Za-z0-9]+",
          "_"
        ) %>%
        stringr::str_replace_all(
          "^_|_$",
          ""
        )
      
      out_prefix <- paste0(
        safe_label,
        "_BBN_UPPL_parental_vs_mp1"
      )
      
    } else {
      
      # Remove any supplied file extension.
      out_prefix <- tools::file_path_sans_ext(
        out_file
      )
    }
    
    # Define and create the output directory.
    out_dir <- here::here(
      "results",
      "plots"
    )
    
    dir.create(
      out_dir,
      recursive = TRUE,
      showWarnings = FALSE
    )
    
    # Save the plot as a high-resolution PNG.
    ggplot2::ggsave(
      filename = file.path(
        out_dir,
        paste0(out_prefix, ".png")
      ),
      plot = activity_plot,
      width = 7,
      height = 4,
      dpi = 300
    )
  }
  
  # Return the plot and calculated means.
  return(
    list(
      activity_plot = activity_plot,
      means = mean_df
    )
  )
}
```

# **Score PPARG Activity in Mouse Samples Using the Public Rosi Signature**

```{r}
# get pparg genes found in rnaseq data
genes_use_pparg <- intersect(umuc9_rosi_upregulated_genes, rownames(humanized_mouse_expr))

#get z score
pparg_gene_z <- t(scale(t(humanized_mouse_expr[genes_use_pparg, , drop = FALSE])))
pparg_score <- colMeans(pparg_gene_z, na.rm = TRUE)

umuc9_pparg_score_df <- format_score(pparg_score, signature_label = "PPARG")

plot_signature_activity(
  score_df = umuc9_pparg_score_df,
  signature_label = "PPARG",
  x_axis_group = "parental",
  stratify_group = "cell_line",
  plot_title="PPARG Activity from UMUC9 Significant Genes",
  out_file="umuc9_public_pparg_activity.png"
)

plot_signature_bbn_vs_uppl(umuc9_pparg_score_df,
                           signature_label = "PPARG",
                           parental_filter = "Parental",
                           # color_group = "cell_model",
                           plot_title = "PPARG (UMUC9) Activity of Parentals",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#0072B2",
                             "UPPL" = "#D55E00"
                           ))

plot_signature_bbn_vs_uppl(umuc9_pparg_score_df,
                           signature_label = "PPARG",
                           parental_filter = "MP1",
                           # color_group = "cell_model",
                           plot_title = "PPARG (UMUC9) Activity of MP1",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#1E3A8A",
                             "UPPL" = "#F5A623"
                           ))

plot_signature_parental_vs_mp1(
  score_df = umuc9_pparg_score_df,
  signature_label = "PPARG",
  cell_model_filter = "BBN",
  condition_cols = c(
    "Parental" = "#2F855A",
    "MP1" = "#F6ADCD"
  ),
  plot_title = "PPARG (UMUC9) Activity of BBNs"
)

plot_signature_parental_vs_mp1(
  score_df = umuc9_pparg_score_df,
  signature_label = "PPARG",
  cell_model_filter = "UPPL",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "PPARG (UMUC9) Activity of UPPLs"
)

plot_signature_parental_vs_mp1_combined(
  umuc9_pparg_score_df, 
  signature_label="PPARG",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ))
```

**Score GATA3 Pathway Activity in Mouse Samples**

```{r}
genes_use_gata3 <- intersect(gata3_sig, rownames(humanized_mouse_expr))

gata3_gene_z <- t(scale(t(humanized_mouse_expr[genes_use_gata3, , drop = FALSE])))
gata3_score <- colMeans(gata3_gene_z, na.rm = TRUE)


gata3_scores_df <- format_score(gata3_score, signature_label = "GATA3")

plot_signature_activity(
  score_df = gata3_scores_df,
  signature_label = "GATA3",
  plot_title="GATA3 Activity",
  out_file="GATA3_activity.png"
)

plot_signature_bbn_vs_uppl(gata3_scores_df,
                           signature_label = "GATA3",
                           parental_filter = "Parental",
                           plot_title = "GATA3 Activity of Parentals",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#0072B2",
                             "UPPL" = "#D55E00"
                           ))

plot_signature_bbn_vs_uppl(gata3_scores_df,
                           signature_label = "GATA3",
                           parental_filter = "MP1",
                           plot_title = "GATA3 Activity of MP1",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#1E3A8A",
                             "UPPL" = "#F5A623"
                           ))

plot_signature_parental_vs_mp1_combined(
  gata3_scores_df, 
  signature_label="GATA3",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ))
```

# **Identify In-House Rosiglitazone-Responsive Genes**

- u1 = umuc1
- u6 = umuc6
- u14 = umuc14
- rt = rt112

```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)

#load in tpm data and convert to log2(tpm+1)
tpm_umuc_rt_rosi_tpm <- read.delim("~/Documents/GitHub/mouse-models/BASE47/data/rosi_counts_tpm_gene_id.tsv", row.names=1)
tpm_umuc_rt_rosi_tpm <- tpm_umuc_rt_rosi_tpm[, !grepl("_N_", names(tpm_umuc_rt_rosi_tpm))]
tpm_umuc_rt_rosi_log2 <- log2(tpm_umuc_rt_rosi_tpm + 1)

#get ensembl id without . 
umuc_rt_rosi_ensembl_genes <- sub("\\.\\d+$", "", rownames(tpm_umuc_rt_rosi_log2))

#get human gene names mapping
umuc_rt112_gene_symbols <- suppressMessages(
  mapIds(
    org.Hs.eg.db,
    keys = umuc_rt_rosi_ensembl_genes,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
)

# Automatically label:
#   rosi-treatment: 1
#   control: 2
rosi_treatment_labels <- ifelse(grepl("_R_", colnames(tpm_umuc_rt_rosi_log2)), 1,
            ifelse(grepl("_U_", colnames(tpm_umuc_rt_rosi_log2)), 2, NA))

stopifnot(!any(is.na(rosi_treatment_labels)))
stopifnot(length(rosi_treatment_labels) == ncol(tpm_umuc_rt_rosi_log2))

sam_data <- list(
  x = as.matrix(tpm_umuc_rt_rosi_log2),
  y = rosi_treatment_labels,
  geneid = rownames(tpm_umuc_rt_rosi_log2),
  genenames = umuc_rt112_gene_symbols,
  logged2 = TRUE
)

samr_obj <- quiet(
  samr(data=sam_data, resp.type = "Two class unpaired", nperms = 100)
)
delta_table <- quiet(
  samr.compute.delta.table(samr_obj)
  )


# Choose a delta where median FDR < 0.05
# -b/c delta != FDR
delta_df <- as.data.frame(delta_table)
chosen_delta <- delta_df[delta_df[, "median FDR"] < 0.05, "delta"][1]

# get significant genes in rosi vs ctrl
umuc_rt112_siggenes <- samr.compute.siggenes.table(
  samr.obj = samr_obj,
  del = chosen_delta,
  data = sam_data,
  delta.table = delta_table
)
```

# **Score Samples Using In-House Rosi-Responsive Signature**

```{r}
umuc_rt112_pparg_signature <- umuc_rt112_siggenes$genes.up[, "Gene ID"]

umuc_rt112_pparg_genes_in_mouse_expr <- intersect(umuc_rt112_pparg_signature, rownames(humanized_mouse_expr))

#get z score
umuc_rt112_pparg_gene_zscores <- t(scale(t(humanized_mouse_expr[umuc_rt112_pparg_genes_in_mouse_expr, , drop = FALSE])))
umuc_rt112_pparg_scores <- colMeans(umuc_rt112_pparg_gene_zscores, na.rm = TRUE)

umuc_rt112_pparg_scores_df <- format_score(umuc_rt112_pparg_scores, signature_label = "PPARG")

plot_signature_activity(
  score_df = umuc_rt112_pparg_scores_df,
  signature_label = "PPARG",
  plot_title="PPARG Activity from UMUCs+RT112 Significant Genes",
  out_file="umuc_rt112_pparg_activity.png"
)

plot_signature_bbn_vs_uppl(umuc_rt112_pparg_scores_df,
                           signature_label = "PPARG",
                           parental_filter = "Parental",
                           color_group = "cell_model",
                           plot_title = "PPARG (UMUCs+RT112) Activity of Parentals",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#0072B2",
                             "UPPL" = "#D55E00"
                           ))

plot_signature_bbn_vs_uppl(umuc_rt112_pparg_scores_df,
                           signature_label = "PPARG",
                           parental_filter = "MP1",
                           color_group = "cell_model",
                           plot_title = "PPARG (UMUCs+RT112) Activity of MP1",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#1E3A8A",
                             "UPPL" = "#F5A623"
                           ))

plot_signature_parental_vs_mp1(
  score_df = umuc_rt112_pparg_scores_df,
  signature_label = "PPARG",
  cell_model_filter = "BBN",
  condition_cols = c(
    "Parental" = "#2F855A",
    "MP1" = "#F6ADCD"
  ),
  plot_title = "PPARG (UMUCs+RT112) Activity of BBNs"
)

plot_signature_parental_vs_mp1(
  score_df = umuc_rt112_pparg_scores_df,
  signature_label = "PPARG",
  cell_model_filter = "UPPL",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "PPARG (UMUCs+RT112) Activity of UPPLs"
)
```

Run SAM on 4 lines separately:

```{r}
# Cell-line identifiers in the sample names
cell_lines <- c(
  umuc1  = "u1",
  umuc6  = "u6",
  umuc14 = "u4",
  rt112  = "r1"
)

run_cell_line_sam <- function(cell_line_id,
                              expression_data,
                              gene_symbols,
                              fdr_cutoff = 0.05,
                              nperms = 100) {
  
  # Match the cell-line ID as a complete underscore-separated token
  # This prevents "u1" from matching "u14"
  cell_line_pattern <- paste0("(^|_)", cell_line_id, "(_|$)")
  
  keep_samples <- grepl(
    cell_line_pattern,
    colnames(expression_data),
    ignore.case = TRUE
  )
  
  cell_line_expression <- expression_data[, keep_samples, drop = FALSE]
  
  if (ncol(cell_line_expression) == 0) {
    stop("No samples found for cell line: ", cell_line_id)
  }
  
  # Rosi treatment = 1; untreated control = 2
  treatment_labels <- ifelse(
    grepl("_R_", colnames(cell_line_expression)),
    1,
    ifelse(
      grepl("_U_", colnames(cell_line_expression)),
      2,
      NA
    )
  )
  
  if (anyNA(treatment_labels)) {
    stop(
      "Could not assign treatment labels for: ",
      paste(colnames(cell_line_expression)[is.na(treatment_labels)],
            collapse = ", ")
    )
  }
  
  if (!all(c(1, 2) %in% treatment_labels)) {
    stop(
      "Cell line ", cell_line_id,
      " does not contain both rosi and control samples."
    )
  }
  
  sam_data <- list(
    x = as.matrix(cell_line_expression),
    y = treatment_labels,
    geneid = rownames(cell_line_expression),
    genenames = gene_symbols,
    logged2 = TRUE
  )
  
  samr_obj <- quiet(
    samr(
      data = sam_data,
      resp.type = "Two class unpaired",
      nperms = nperms
    )
  )
  
  delta_table <- quiet(
    samr.compute.delta.table(samr_obj)
  )
  
  delta_df <- as.data.frame(delta_table)
  
  eligible_deltas <- delta_df[
    !is.na(delta_df[["median FDR"]]) &
      delta_df[["median FDR"]] < fdr_cutoff,
    "delta"
  ]
  
  if (length(eligible_deltas) == 0) {
    warning(
      "No delta produced a median FDR below ",
      fdr_cutoff,
      " for cell line ",
      cell_line_id
    )
    
    return(list(
      siggenes = NULL,
      chosen_delta = NA_real_,
      delta_table = delta_table,
      samr_obj = samr_obj,
      samples = colnames(cell_line_expression)
    ))
  }
  
  chosen_delta <- eligible_deltas[1]
  
  siggenes <- samr.compute.siggenes.table(
    samr.obj = samr_obj,
    del = chosen_delta,
    data = sam_data,
    delta.table = delta_table
  )
  
  list(
    siggenes = siggenes,
    chosen_delta = chosen_delta,
    delta_table = delta_table,
    samr_obj = samr_obj,
    samples = colnames(cell_line_expression)
  )
}

# Run a separate SAM analysis for each cell line
rosi_sam_results <- lapply(
  cell_lines,
  run_cell_line_sam,
  expression_data = tpm_umuc_rt_rosi_log2,
  gene_symbols = umuc_rt112_gene_symbols,
  fdr_cutoff = 0.05,
  nperms = 100
)

# Four significant-gene table objects
umuc1_siggenes  <- rosi_sam_results$umuc1$siggenes$genes.up[, "Gene Name"]
umuc6_siggenes  <- rosi_sam_results$umuc6$siggenes$genes.up[, "Gene Name"]
umuc14_siggenes <- rosi_sam_results$umuc14$siggenes$genes.up[, "Gene Name"]
rt112_siggenes  <- rosi_sam_results$rt112$siggenes$genes.up[, "Gene Name"]
```

```{r}
# install.packages("ComplexUpset")

library(ComplexUpset)
library(ggplot2)

upregulated_gene_sets <- list(
  UMUC1  = unique(na.omit(umuc1_siggenes)),
  UMUC6  = unique(na.omit(umuc6_siggenes)),
  UMUC14 = unique(na.omit(umuc14_siggenes)),
  RT112  = unique(na.omit(rt112_siggenes)),
  UMUC9 = unique(na.omit(umuc9_rosi_upregulated_genes))
)

all_genes <- sort(unique(unlist(upregulated_gene_sets)))

upset_data <- data.frame(
  gene = all_genes,
  UMUC1 = all_genes %in% upregulated_gene_sets$UMUC1,
  UMUC6 = all_genes %in% upregulated_gene_sets$UMUC6,
  UMUC14 = all_genes %in% upregulated_gene_sets$UMUC14,
  RT112 = all_genes %in% upregulated_gene_sets$RT112,
  UMUC9 = all_genes %in% upregulated_gene_sets$UMUC9
)

upset_plot <- ComplexUpset::upset(
  upset_data,
  intersect = c("UMUC1", "UMUC6", "UMUC14", "RT112", "UMUC9"),
  name = "Cell line",
  min_size = 0,
  sort_intersections_by = c("degree", "cardinality"),
  base_annotations = list(
    "Intersection size" =
      intersection_size(counts = TRUE)
  ),
  set_sizes = (
    upset_set_size() +
      ylab("Upregulated genes")
  )
) +
  ggtitle("Overlap of significantly upregulated genes")

upset_plot

ggsave(
  here("results","plots","rosi_upregulated_genes_upset_plot.png"),
  upset_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

Overlapping Significantly Upregulated Genes between UMUC9 and UMUCs+RT112

```{r}
library(ggforce)
library(ggrepel)


umuc_rt112_pparg_signature

genes_use_pparg
highlight_genes <- c("TP53", "BRCA1", "NECTIN4", "FABP4", "GATA3", "KRT6", "KRT14", "KRT20", "PPARG")

umuc9_only <- setdiff(genes_use_pparg, umuc_rt112_pparg_signature)
umuc_rt112_only <- setdiff(umuc_rt112_pparg_signature, genes_use_pparg)
both_umuc9_umuc_rt112   <- intersect(genes_use_pparg, umuc_rt112_only)

counts <- data.frame(
  region = c("UMUC9 only", "Both", "UMUC+RT112 only"),
  x = c(-0.8, 0, 0.8),
  y = c(0, 0, 0),
  label = c(length(umuc9_only), length(both_umuc9_umuc_rt112), length(umuc_rt112_only))
)

highlight_df <- data.frame(gene = highlight_genes) %>%
  filter(gene %in% union(umuc9_only, umuc_rt112_only)) %>%
  mutate(
    region = case_when(
      gene %in% both_umuc9_umuc_rt112 ~ "Both",
      gene %in% umuc9_only ~ "A only",
      gene %in% umuc_rt112_only ~ "B only"
    ),
    x = case_when(
      region == "A only" ~ -0.8,
      region == "Both" ~ 0,
      region == "B only" ~ 0.8
    ),
    y = row_number() * 0.15
  )


ggplot() +
  geom_circle(
    aes(x0 = -0.5, y0 = 0, r = 1),
    fill = "skyblue",
    alpha = 0.35,
    color = "black"
  ) +
  geom_circle(
    aes(x0 = 0.5, y0 = 0, r = 1),
    fill = "salmon",
    alpha = 0.35,
    color = "black"
  ) +
  geom_text(
    data = counts,
    aes(x = x, y = y, label = label),
    size = 7
  ) +
  geom_text_repel(
    data = highlight_df,
    aes(x = x, y = y, label = gene),
    size = 4,
    fontface = "bold",
    box.padding = 0.4
  ) +
  annotate("text", x = -0.9, y = 1.15, label = "UMUC9", size = 5) +
  annotate("text", x = 0.9, y = 1.15, label = "UMUC+RT112", size = 5) +
  coord_fixed() +
  theme_void()
```
