---
title: "GATA3/PPARG Scoring"
format: html
editor: visual
---

Folder structure: GATA3_PPARG_Scoring/ \|-- data/ \|-- results/ \| `-- plots/`-- scripts/ \`-- run_gata3_pparg_files/ ---

## Setup

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(here)
# here::i_am("README.Rmd")

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("GEOquery", "impute"))
library(samr)
library(GEOquery)
library(limma)
library(tidyverse)

# =========
# Functions
# =========

#Convert to 90th percentile of log(tpm+1)
cpm_to_uq_log2 <- function(cpm_matrix) {
  uq <- apply(cpm_matrix, 2, function(x) quantile(x[x > 0], 0.75))
  scaled <- sweep(cpm_matrix, 2, uq, "/") * 1000
  log2(scaled + 1)
}

# Suppress permutations output
quiet <- function(expr) {
  suppressMessages(suppressWarnings({
    capture.output(result <- eval.parent(substitute(expr)))
    result
  }))
}
```

## Derive Public Rosiglitazone-Induced PPARγ Pathway Signature from GSE47993

1.  Downloads public FABP4-modulation GEO dataset GSE47993
2.  Subset to only UMUC9 control and rosi-treated samples
    1.  Uses rosiglitazone treatment as a PPARγ activation condition
3.  Map Illumina Probe IDs to gene symbols
4.  Collapse multiple probes mapping to the same gene by averaging expression values
5.  Uses SAM to identify genes significantly upregulated after rosi treatment

```{r}
gse47993_dataset <- getGEO("GSE47993", GSEMatrix = TRUE)[[1]]
gse47993_expr <- exprs(gse47993_dataset)

umuc9_rosi_expr <- gse47993_expr[, c("GSM1164283","GSM1164284","GSM1164285",  # control samples
                       "GSM1164286","GSM1164287","GSM1164288") # rosi treated samples
]

# Get gene names from illumina probe IDs (they used IlluminaHT-12 V4.0 expression beadchip)
gse47993_annotation <- fData(gse47993_dataset)
probe_gene_symbols <- gse47993_annotation[rownames(umuc9_rosi_expr), "Symbol"]

# keep probes with valid gene symbols
umuc9_expr_with_symbols <- umuc9_rosi_expr[!is.na(probe_gene_symbols) & probe_gene_symbols != "", ]
probe_gene_symbols <- probe_gene_symbols[!is.na(probe_gene_symbols) & probe_gene_symbols != ""]
# Collapse multiple probes per gene
umuc9_gene_expr <- avereps(umuc9_expr_with_symbols, ID = probe_gene_symbols) # now have gene expression x sample matrix
rosi_group_labels <- c(1,1,1,2,2,2) #1=ctrl, #2=rosi-treated

umuc9_sam_input <- list(
  x = umuc9_gene_expr, 
  y = rosi_group_labels,
  geneid = rownames(umuc9_gene_expr),
  genenames = rownames(umuc9_gene_expr),
  logged2 = TRUE
)

umuc9_sam_fit <- quiet(
  samr(
    data=umuc9_sam_input, resp.type = "Two class unpaired", nperms = 100)
)
delta_table <- quiet(
  samr.compute.delta.table(umuc9_sam_fit)
)


# Choose a delta where median FDR < 0.05
# -b/c delta != FDR
sam_delta_results <- as.data.frame(delta_table)
fdr05_delta <- sam_delta_results[sam_delta_results[, "median FDR"] < 0.05, "delta"][1]

# get significant genes in rosi vs ctrl
umuc9_rosi_siggenes <- samr.compute.siggenes.table(
  samr.obj = umuc9_sam_fit,
  del = fdr05_delta,
  data = umuc9_sam_input,
  delta.table = delta_table
)
```

**Extract Rosi-Induced PPARG Target Genes**

- Saves significantly upregulated rosiglitazone-responsive genes as PPARG activity signature
- These genes represent downstream transcriptional changes associated with PPARG activation

```{r}
# Significant upregulated genes AKA genes induced by rosi (downstream most likely)
umuc9_rosi_upregulated_genes <- umuc9_rosi_siggenes$genes.up[, "Gene Name"]
head(umuc9_rosi_upregulated_genes)

#number of genes upregulated
umuc9_rosi_siggenes$ngenes.up
```

**Define Curated GATA3 Pathway Signature**

```{r}
gata3_sig <- c(
  "ADCY1", "GATA3", "GNAS", "IL13", "IL1A", "IL4", "IL5",
  "JUNB", "MAF", "MAP2K3", "MAP2K6", "MAPK14", "NFATC1",
  "PRKACB", "PRKACG", "PRKAR1A", "PRKAR1B", "PRKAR2A", "PRKAR2B"
)
```

**Load, Normalize, and Map Mouse RNA-seq Data to Human Gene Symbols**

- Loads parental and MP1 mouse RNA-seq expression matrix
- Removes count, gene ID, and gene biotype columns not needed for scoring
- Keeps one row per mouse gene symbol
- Applies upper-quartile log2 normalization (USES CPM, NOT TPM)
- Converts mouse gene symbols to human orthologs so they can be compared with human PPARG and GATA3 signatures
- Collapses duplicate human ortholog mappings by averaging expression values

```{r}
mouse_cpm_expr <- read.csv(here("data", "parental_mp1_K8HVZP-expression-matrix.csv"), sep=",", header=T)
mouse_cpm_expr <- mouse_cpm_expr[, !grepl("count|^gene_biotype|gene_id", colnames(mouse_cpm_expr)    )    ] #remove count, gene_biotype, gene_id columns
mouse_cpm_expr <- mouse_cpm_expr %>% #remove duplicate reads
  mutate(row_avg = rowMeans(pick(where(is.numeric)), na.rm = TRUE)) %>% #make new column for avg expression across all samples
  arrange(desc(row_avg)) %>%
  distinct(gene_name, .keep_all=TRUE) %>% #keep only read w/ highest avg read across all samples
  dplyr::select(-row_avg) #remove temp column
rownames(mouse_cpm_expr) <- mouse_cpm_expr$gene_name
mouse_cpm_expr$gene_name <- NULL

# Convert cpm to upper quartile log2(cpm+1)
# IMPORTANT: WE USED CPM BUT THEY USED TPM
# SINCE WE ONLY WERE GIVEN CPM AND I DIDN'T WANT TO PROCESS FOR TPM
mouse_uqlog2_expr <- cpm_to_uq_log2(mouse_cpm_expr)

#get mouse symbols from our samples
mouse_gene_symbols <- unique(rownames(mouse_uqlog2_expr))

# get map of mouse:human genes for conversion
mouse_to_human_orthologs <- babelgene::orthologs(
  genes = mouse_gene_symbols,
  species = "mouse",
  human = FALSE,
  top=TRUE
)

#Make new column for human gene ortholog
mouse_uqlog2_expr$human_symbol <- mouse_to_human_orthologs$human_symbol[
  match(rownames(mouse_uqlog2_expr), mouse_to_human_orthologs$symbol)
]

#keep only reads that have human ortholog
sample_cols <- setdiff(colnames(mouse_uqlog2_expr), "human_symbol")

humanized_mouse_expr <- mouse_uqlog2_expr %>%
  filter(!is.na(human_symbol), human_symbol != "") %>% #remove NAs
  group_by(human_symbol) %>%
  summarise(across(all_of(sample_cols), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  column_to_rownames("human_symbol")
```

**Signature Activity Plotting Function**

- Builds plotting function that can be reused for PPARG, GATA3, or other gene signatures
- Converts signature scores into organized plotting table
- Annotates each sample by model, such as `BBN963` or `UPPL1541`.
- Annotates each sample by subtype (parental or MP1)
- Creates boxplot with jittered sample-level points
- Saves plot to `results/plots/` and also returns plot object

```{r}
format_score <- function(score_vec, signature_label = NULL) {
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single string.")
  }
  
  if (is.null(names(score_vec))) {
    stop("score_vec must be a named vector.")
  }
  
  score_col <- paste0(signature_label, "_z_score")
  sample_names <- names(score_vec)
  
  score_df <- data.frame(
    cell_line = sub("_.*", "", sample_names),
    parental = ifelse(grepl("_PARENTAL_", sample_names), "Parental", "MP1"),
    score_vec,
    check.names = FALSE
  )
  
  names(score_df)[names(score_df) == "score_vec"] <- score_col
  
  return(score_df)
}

plot_signature_activity <- function(score_df,
                                    signature_label = NULL,
                                    x_group = "cell_line",
                                    color_group = "parental",
                                    plot_title = NULL,
                                    y_axis_label = NULL,
                                    out_file = NULL,
                                    save_plot = TRUE) {
  
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single string.")
  }
  
  score_col <- paste0(signature_label, "_z_score")
  
  if (!score_col %in% colnames(score_df)) {
    stop(paste0("Column '", score_col, "' not found in score_df."))
  }
  
  if (!x_group %in% colnames(score_df)) {
    stop(paste0("Column '", x_group, "' not found in score_df."))
  }
  
  if (!color_group %in% colnames(score_df)) {
    stop(paste0("Column '", color_group, "' not found in score_df."))
  }
  
  if (is.null(plot_title)) {
    plot_title <- paste(signature_label, "Activity")
  }
  
  if (is.null(y_axis_label)) {
    y_axis_label <- paste(signature_label, "Signature Z-Score")
  }
  
  score_df <- score_df %>%
    mutate(
      across(all_of(c(x_group, color_group)), as.factor),
      "{color_group}" := factor(
        .data[[color_group]],
        levels = c("Parental", "MP1")
      )
    ) %>%
    filter(
      !is.na(.data[[x_group]]),
      !is.na(.data[[color_group]]),
      !is.na(.data[[score_col]])
    )
  
  condition_cols <- c(
    "Parental" = "#18AEB5",
    "MP1" = "#F05A50"
  )
  
  model_cols <- c(
    "BBN963" = "#0072B2",
    "UPPL1541" = "#D55E00"
  )
  
  grouped_dot_plot <- ggplot(
    score_df,
    aes(
      x = .data[[color_group]],
      y = .data[[score_col]],
      color = .data[[color_group]]
    )
  ) +
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    geom_jitter(
      width = 0.08,
      height = 0,
      size = 2.8,
      alpha = 0.9
    ) +
    stat_summary(
      fun = mean,
      geom = "point",
      shape = 95,
      size = 10,
      linewidth = 1.2
    ) +
    facet_wrap(vars(.data[[x_group]]), nrow = 1) +
    scale_color_manual(values = condition_cols, drop = FALSE) +
    labs(
      title = plot_title,
      x = NULL,
      y = y_axis_label,
      color = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )
  
  mean_df <- score_df %>%
    group_by(
      .data[[x_group]],
      .data[[color_group]]
    ) %>%
    summarise(
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  interaction_plot <- ggplot() +
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    geom_jitter(
      data = score_df,
      aes(
        x = .data[[color_group]],
        y = .data[[score_col]],
        color = .data[[x_group]]
      ),
      width = 0.08,
      height = 0,
      size = 2.4,
      alpha = 0.45
    ) +
    geom_line(
      data = mean_df,
      aes(
        x = .data[[color_group]],
        y = mean_score,
        group = .data[[x_group]],
        color = .data[[x_group]]
      ),
      linewidth = 1.1
    ) +
    geom_point(
      data = mean_df,
      aes(
        x = .data[[color_group]],
        y = mean_score,
        color = .data[[x_group]]
      ),
      size = 3.5
    ) +
    scale_color_manual(values = model_cols) +
    labs(
      title = paste(plot_title, "\nInteraction View"),
      x = NULL,
      y = y_axis_label,
      color = "Cell line"
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "right"
    )
  
  if (save_plot) {
    
    if (is.null(out_file)) {
      safe_label <- signature_label %>%
        str_replace_all("[^A-Za-z0-9]+", "_") %>%
        str_replace_all("^_|_$", "")
      
      out_prefix <- paste0(safe_label, "_activity")
    } else {
      out_prefix <- tools::file_path_sans_ext(out_file)
    }
    
    out_dir <- here::here("results", "plots")
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    
    ggsave(
      filename = file.path(out_dir, paste0(out_prefix, "_grouped_dot_plot.png")),
      plot = grouped_dot_plot,
      width = 6,
      height = 4,
      dpi = 300
    )
    
    ggsave(
      filename = file.path(out_dir, paste0(out_prefix, "_interaction_plot.png")),
      plot = interaction_plot,
      width = 5,
      height = 4,
      dpi = 300
    )
  }
  
  return(list(
    grouped_dot_plot = grouped_dot_plot,
    interaction_plot = interaction_plot,
    means = mean_df
  ))
}

plot_signature_activity <- function(score_df,
                                    signature_label = NULL,
                                    x_group = "cell_line",
                                    color_group = "parental",
                                    plot_title = NULL,
                                    y_axis_label = NULL,
                                    out_file = NULL,
                                    save_plot = TRUE) {
  
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single string.")
  }
  
  score_col <- paste0(signature_label, "_z_score")
  
  if (!score_col %in% colnames(score_df)) {
    stop(paste0("Column '", score_col, "' not found in score_df."))
  }
  
  if (!x_group %in% colnames(score_df)) {
    stop(paste0("Column '", x_group, "' not found in score_df."))
  }
  
  if (!color_group %in% colnames(score_df)) {
    stop(paste0("Column '", color_group, "' not found in score_df."))
  }
  
  if (is.null(plot_title)) {
    plot_title <- paste(signature_label, "Activity")
  }
  
  if (is.null(y_axis_label)) {
    y_axis_label <- paste(signature_label, "Signature Z-Score")
  }
  
  score_df <- score_df %>%
    mutate(
      across(all_of(c(x_group, color_group)), as.factor),
      "{color_group}" := factor(
        .data[[color_group]],
        levels = c("Parental", "MP1")
      )
    ) %>%
    filter(
      !is.na(.data[[x_group]]),
      !is.na(.data[[color_group]]),
      !is.na(.data[[score_col]])
    )
  
  condition_cols <- c(
    "Parental" = "#18AEB5",
    "MP1" = "#F05A50"
  )
  
  model_cols <- c(
    "BBN963" = "#0072B2",
    "UPPL1541" = "#D55E00"
  )
  
  grouped_dot_plot <- ggplot(
    score_df,
    aes(
      x = .data[[color_group]],
      y = .data[[score_col]],
      color = .data[[color_group]]
    )
  ) +
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    geom_jitter(
      width = 0.08,
      height = 0,
      size = 2.8,
      alpha = 0.9
    ) +
    stat_summary(
      fun = mean,
      geom = "point",
      shape = 95,
      size = 10,
      linewidth = 1.2
    ) +
    facet_wrap(vars(.data[[x_group]]), nrow = 1) +
    scale_color_manual(values = condition_cols, drop = FALSE) +
    labs(
      title = plot_title,
      x = NULL,
      y = y_axis_label,
      color = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )
  
  mean_df <- score_df %>%
    group_by(
      .data[[x_group]],
      .data[[color_group]]
    ) %>%
    summarise(
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  interaction_plot <- ggplot() +
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    geom_jitter(
      data = score_df,
      aes(
        x = .data[[color_group]],
        y = .data[[score_col]],
        color = .data[[x_group]]
      ),
      width = 0.08,
      height = 0,
      size = 2.4,
      alpha = 0.45
    ) +
    geom_line(
      data = mean_df,
      aes(
        x = .data[[color_group]],
        y = mean_score,
        group = .data[[x_group]],
        color = .data[[x_group]]
      ),
      linewidth = 1.1
    ) +
    geom_point(
      data = mean_df,
      aes(
        x = .data[[color_group]],
        y = mean_score,
        color = .data[[x_group]]
      ),
      size = 3.5
    ) +
    scale_color_manual(values = model_cols) +
    labs(
      title = paste(plot_title, "\nInteraction View"),
      x = NULL,
      y = y_axis_label,
      color = "Cell line"
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "right"
    )
  
  if (save_plot) {
    
    if (is.null(out_file)) {
      safe_label <- signature_label %>%
        str_replace_all("[^A-Za-z0-9]+", "_") %>%
        str_replace_all("^_|_$", "")
      
      out_prefix <- paste0(safe_label, "_activity")
    } else {
      out_prefix <- tools::file_path_sans_ext(out_file)
    }
    
    out_dir <- here::here("results", "plots")
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    
    ggsave(
      filename = file.path(out_dir, paste0(out_prefix, "_grouped_dot_plot.png")),
      plot = grouped_dot_plot,
      width = 6,
      height = 4,
      dpi = 300
    )
    
    ggsave(
      filename = file.path(out_dir, paste0(out_prefix, "_interaction_plot.png")),
      plot = interaction_plot,
      width = 5,
      height = 4,
      dpi = 300
    )
  }
  
  return(list(
    grouped_dot_plot = grouped_dot_plot,
    interaction_plot = interaction_plot,
    means = mean_df
  ))
}

plot_signature_bbn_vs_uppl <- function(score_df,
                                       signature_label = NULL,
                                       parental_filter = "Parental",
                                       plot_title = NULL,
                                       out_file = NULL,
                                       save_plot = TRUE,
                                       model_cols = c(
                                         "BBN" = "#0072B2",
                                         "UPPL" = "#D55E00"
                                       )) {
  
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single string.")
  }
  
  score_col <- paste0(signature_label, "_z_score")
  
  if (!score_col %in% colnames(score_df)) {
    stop(paste0("Column '", score_col, "' not found in score_df."))
  }
  
  if (!"cell_line" %in% colnames(score_df)) {
    stop("Column 'cell_line' not found in score_df.")
  }
  
  if (!"parental" %in% colnames(score_df)) {
    stop("Column 'parental' not found in score_df.")
  }
  
  if (is.null(plot_title)) {
    plot_title <- paste0(signature_label, " Activity: ", parental_filter, " Only")
  }
  
  if (is.null(y_axis_label)) {
    y_axis_label <- paste(signature_label, "Signature Z-Score")
  }
  
  score_df <- score_df %>%
    filter(parental == parental_filter) %>%
    mutate(
      cell_model = case_when(
        grepl("^BBN", cell_line) ~ "BBN",
        grepl("^UPPL", cell_line) ~ "UPPL",
        TRUE ~ NA_character_
      ),
      cell_model = factor(cell_model, levels = c("BBN", "UPPL"))
    ) %>%
    filter(
      !is.na(cell_model),
      !is.na(.data[[score_col]])
    )
  
  activity_plot <- ggplot(
    score_df,
    aes(
      x = cell_model,
      y = .data[[score_col]],
      color = cell_model
    )
  ) +
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    geom_jitter(
      width = 0.08,
      height = 0,
      size = 2.8,
      alpha = 0.9
    ) +
    stat_summary(
      fun = mean,
      geom = "point",
      shape = 95,
      size = 10,
      linewidth = 1.2
    ) +
    scale_color_manual(values = model_cols, drop = FALSE) +
    labs(
      title = plot_title,
      x = NULL,
      y = y_axis_label,
      color = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "right"
    )
  
  mean_df <- score_df %>%
    group_by(cell_model) %>%
    summarise(
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  if (save_plot) {
    
    if (is.null(out_file)) {
      safe_label <- signature_label %>%
        str_replace_all("[^A-Za-z0-9]+", "_") %>%
        str_replace_all("^_|_$", "")
      
      safe_condition <- parental_filter %>%
        str_replace_all("[^A-Za-z0-9]+", "_") %>%
        str_replace_all("^_|_$", "")
      
      out_prefix <- paste0(safe_label, "_", safe_condition, "_bbn_vs_uppl")
    } else {
      out_prefix <- tools::file_path_sans_ext(out_file)
    }
    
    out_dir <- here::here("results", "plots")
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    
    ggsave(
      filename = file.path(out_dir, paste0(out_prefix, ".png")),
      plot = activity_plot,
      width = 4.5,
      height = 4,
      dpi = 300
    )
  }
  
  return(list(
    activity_plot = activity_plot,
    means = mean_df
  ))
}

plot_signature_parental_vs_mp1 <- function(score_df,
                                           signature_label = NULL,
                                           cell_model_filter = "BBN",
                                           x_group = "parental",
                                           color_group = "parental",
                                           plot_title = NULL,
                                           out_file = NULL,
                                           save_plot = TRUE,
                                           condition_cols = c(
                                             "Parental" = "#18AEB5",
                                             "MP1" = "#F05A50"
                                           )) {
  
  if (!is.character(signature_label) ||
      length(signature_label) != 1 ||
      is.na(signature_label) ||
      signature_label == "") {
    stop("signature_label must be a single string.")
  }
  
  score_col <- paste0(signature_label, "_z_score")
  
  if (!score_col %in% colnames(score_df)) {
    stop(paste0("Column '", score_col, "' not found in score_df."))
  }
  
  if (!"cell_line" %in% colnames(score_df)) {
    stop("Column 'cell_line' not found in score_df.")
  }
  
  if (!"parental" %in% colnames(score_df)) {
    stop("Column 'parental' not found in score_df.")
  }
  
  if (is.null(plot_title)) {
    plot_title <- paste0(signature_label, " Activity: ", cell_model_filter, " Only")
  }
    y_axis_label <- paste(signature_label, "Signature Z-Score")
  
  score_df <- score_df %>%
    mutate(
      cell_model = case_when(
        grepl("^BBN", cell_line) ~ "BBN",
        grepl("^UPPL", cell_line) ~ "UPPL",
        TRUE ~ NA_character_
      ),
      parental = factor(parental, levels = c("Parental", "MP1"))
    ) %>%
    filter(
      cell_model == cell_model_filter,
      !is.na(parental),
      !is.na(.data[[score_col]])
    )
  
  activity_plot <- ggplot(
    score_df,
    aes(
      x = .data[[x_group]],
      y = .data[[score_col]],
      color = .data[[color_group]]
    )
  ) +
    geom_hline(
      yintercept = 0,
      linetype = "dotted",
      linewidth = 0.5
    ) +
    geom_jitter(
      width = 0.08,
      height = 0,
      size = 2.8,
      alpha = 0.9
    ) +
    stat_summary(
      fun = mean,
      geom = "point",
      shape = 95,
      size = 10,
      linewidth = 1.2
    ) +
    scale_color_manual(values = condition_cols, drop = FALSE) +
    labs(
      title = plot_title,
      x = NULL,
      y = y_axis_label,
      color = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "right"
    )
  
  mean_df <- score_df %>%
    group_by(.data[[x_group]]) %>%
    summarise(
      mean_score = mean(.data[[score_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  if (save_plot) {
    
    if (is.null(out_file)) {
      safe_label <- signature_label %>%
        str_replace_all("[^A-Za-z0-9]+", "_") %>%
        str_replace_all("^_|_$", "")
      
      out_prefix <- paste0(safe_label, "_", cell_model_filter, "_parental_vs_mp1")
    } else {
      out_prefix <- tools::file_path_sans_ext(out_file)
    }
    
    out_dir <- here::here("results", "plots")
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    
    ggsave(
      filename = file.path(out_dir, paste0(out_prefix, ".png")),
      plot = activity_plot,
      width = 4.5,
      height = 4,
      dpi = 300
    )
  }
  
  return(list(
    activity_plot = activity_plot,
    means = mean_df
  ))
}
```

**Score PPARG Activity in Mouse Samples Using the Public Rosi Signature**

```{r}
# get pparg genes found in rnaseq data
genes_use_pparg <- intersect(umuc9_rosi_upregulated_genes, rownames(humanized_mouse_expr))

#get z score
pparg_gene_z <- t(scale(t(humanized_mouse_expr[genes_use_pparg, , drop = FALSE])))
pparg_score <- colMeans(pparg_gene_z, na.rm = TRUE)

umuc9_pparg_score_df <- format_score(pparg_score, signature_label = "PPARG")

plot_signature_activity(
  score_df = umuc9_pparg_score_df,
  signature_label = "PPARG",
  plot_title="PPARG Activity from UMUC9 Significant Genes",
  out_file="umuc9_public_pparg_activity.png"
)

plot_signature_bbn_vs_uppl(umuc9_pparg_score_df,
                           signature_label = "PPARG",
                           parental_filter = "Parental",
                           color_group = "cell_model",
                           plot_title = "PPARG (UMUC9) Activity of Parentals",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#0072B2",
                             "UPPL" = "#D55E00"
                           ))

plot_signature_bbn_vs_uppl(umuc9_pparg_score_df,
                           signature_label = "PPARG",
                           parental_filter = "MP1",
                           color_group = "cell_model",
                           plot_title = "PPARG (UMUC9) Activity of MP1",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#1E3A8A",
                             "UPPL" = "#F5A623"
                           ))

plot_signature_parental_vs_mp1(
  score_df = umuc9_pparg_score_df,
  signature_label = "PPARG",
  cell_model_filter = "BBN",
  condition_cols = c(
    "Parental" = "#2F855A",
    "MP1" = "#F6ADCD"
  ),
  plot_title = "PPARG (UMUC9) Activity of BBNs"
)

plot_signature_parental_vs_mp1(
  score_df = umuc9_pparg_score_df,
  signature_label = "PPARG",
  cell_model_filter = "UPPL",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "PPARG (UMUC9) Activity of UPPLs"
)
```

**Score GATA3 Pathway Activity in Mouse Samples**

```{r}
genes_use_gata3 <- intersect(gata3_sig, rownames(humanized_mouse_expr))

gata3_gene_z <- t(scale(t(humanized_mouse_expr[genes_use_gata3, , drop = FALSE])))
gata3_score <- colMeans(gata3_gene_z, na.rm = TRUE)


gata3_scores_df <- format_score(gata3_score, signature_label = "GATA3")

plot_signature_activity(
  score_df = gata3_scores_df,
  signature_label = "GATA3",
  plot_title="GATA3 Activity",
  out_file="GATA3_activity.png"
)

plot_signature_bbn_vs_uppl(gata3_scores_df,
                           signature_label = "GATA3",
                           parental_filter = "Parental",
                           color_group = "cell_model",
                           plot_title = "GATA3 Activity of Parentals",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#0072B2",
                             "UPPL" = "#D55E00"
                           ))

plot_signature_bbn_vs_uppl(gata3_scores_df,
                           signature_label = "GATA3",
                           parental_filter = "MP1",
                           color_group = "cell_model",
                           plot_title = "GATA3 Activity of MP1",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#1E3A8A",
                             "UPPL" = "#F5A623"
                           ))

plot_signature_parental_vs_mp1(
  score_df = gata3_scores_df,
  signature_label = "GATA3",
  cell_model_filter = "BBN",
  condition_cols = c(
    "Parental" = "#2F855A",
    "MP1" = "#F6ADCD"
  ),
  plot_title = "GATA3 Activity of BBNs"
)

plot_signature_parental_vs_mp1(
  score_df = gata3_scores_df,
  signature_label = "GATA3",
  cell_model_filter = "UPPL",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "GATA3 Activity of UPPLs"
)
```

**Identify In-House Rosiglitazone-Responsive Genes**

- u1 = umuc1
- u6 = umuc6
- u14 = umuc14
- rt = rt112

```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)

#load in tpm data and convert to log2(tpm+1)
tpm_umuc_rt_rosi_tpm <- read.delim("~/Documents/GitHub/mouse-models/BASE47/data/rosi_counts_tpm_gene_id.tsv", row.names=1)
tpm_umuc_rt_rosi_tpm <- tpm_umuc_rt_rosi_tpm[, !grepl("_N_", names(tpm_umuc_rt_rosi_tpm))]
tpm_umuc_rt_rosi_log2 <- log2(tpm_umuc_rt_rosi_tpm + 1)

#get ensembl id without . 
umuc_rt_rosi_ensembl_genes <- sub("\\.\\d+$", "", rownames(tpm_umuc_rt_rosi_log2))

#get human gene names mapping
umuc_rt112_gene_symbols <- suppressMessages(
  mapIds(
    org.Hs.eg.db,
    keys = umuc_rt_rosi_ensembl_genes,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
)

# Automatically label:
#   rosi-treatment: 1
#   control: 2
rosi_treatment_labels <- ifelse(grepl("_R_", colnames(tpm_umuc_rt_rosi_log2)), 1,
            ifelse(grepl("_U_", colnames(tpm_umuc_rt_rosi_log2)), 2, NA))

stopifnot(!any(is.na(rosi_treatment_labels)))
stopifnot(length(rosi_treatment_labels) == ncol(tpm_umuc_rt_rosi_log2))

sam_data <- list(
  x = as.matrix(tpm_umuc_rt_rosi_log2),
  y = rosi_treatment_labels,
  geneid = rownames(tpm_umuc_rt_rosi_log2),
  genenames = umuc_rt112_gene_symbols,
  logged2 = TRUE
)

samr_obj <- quiet(
  samr(data=sam_data, resp.type = "Two class unpaired", nperms = 100)
)
delta_table <- quiet(
  samr.compute.delta.table(samr_obj)
  )


# Choose a delta where median FDR < 0.05
# -b/c delta != FDR
delta_df <- as.data.frame(delta_table)
chosen_delta <- delta_df[delta_df[, "median FDR"] < 0.05, "delta"][1]

# get significant genes in rosi vs ctrl
umuc_rt112_siggenes <- samr.compute.siggenes.table(
  samr.obj = samr_obj,
  del = chosen_delta,
  data = sam_data,
  delta.table = delta_table
)
```

**Score Samples Using In-House Rosi-Responsive Signature**

```{r}
umuc_rt112_pparg_signature <- umuc_rt112_siggenes$genes.up[, "Gene ID"]

umuc_rt112_pparg_genes_in_mouse_expr <- intersect(umuc_rt112_pparg_signature, rownames(humanized_mouse_expr))

#get z score
umuc_rt112_pparg_gene_zscores <- t(scale(t(humanized_mouse_expr[umuc_rt112_pparg_genes_in_mouse_expr, , drop = FALSE])))
umuc_rt112_pparg_scores <- colMeans(umuc_rt112_pparg_gene_zscores, na.rm = TRUE)

umuc_rt112_pparg_scores_df <- format_score(umuc_rt112_pparg_scores, signature_label = "PPARG")

plot_signature_activity(
  score_df = umuc_rt112_pparg_scores_df,
  signature_label = "PPARG",
  plot_title="PPARG Activity from UMUCs+RT112 Significant Genes",
  out_file="umuc_rt112_pparg_activity.png"
)

plot_signature_bbn_vs_uppl(umuc_rt112_pparg_scores_df,
                           signature_label = "PPARG",
                           parental_filter = "Parental",
                           color_group = "cell_model",
                           plot_title = "PPARG (UMUCs+RT112) Activity of Parentals",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#0072B2",
                             "UPPL" = "#D55E00"
                           ))

plot_signature_bbn_vs_uppl(umuc_rt112_pparg_scores_df,
                           signature_label = "PPARG",
                           parental_filter = "MP1",
                           color_group = "cell_model",
                           plot_title = "PPARG (UMUCs+RT112) Activity of MP1",
                           out_file = NULL,
                           save_plot = TRUE,
                           model_cols = c(
                             "BBN" = "#1E3A8A",
                             "UPPL" = "#F5A623"
                           ))

plot_signature_parental_vs_mp1(
  score_df = umuc_rt112_pparg_scores_df,
  signature_label = "PPARG",
  cell_model_filter = "BBN",
  condition_cols = c(
    "Parental" = "#2F855A",
    "MP1" = "#F6ADCD"
  ),
  plot_title = "PPARG (UMUCs+RT112) Activity of BBNs"
)

plot_signature_parental_vs_mp1(
  score_df = umuc_rt112_pparg_scores_df,
  signature_label = "PPARG",
  cell_model_filter = "UPPL",
  condition_cols = c(
    "Parental" = "#556B2F",
    "MP1" = "#D8A3B0"
  ),
  plot_title = "PPARG (UMUCs+RT112) Activity of UPPLs"
)
```

Overlapping Significantly Upregulated Genes between UMUC9 and UMUCs+RT112

```{r}
library(ggforce)
library(ggrepel)


umuc_rt112_pparg_signature

genes_use_pparg
highlight_genes <- c("TP53", "BRCA1", "NECTIN4", "FABP4")

umuc9_only <- setdiff(genes_use_pparg, umuc_rt112_pparg_signature)
umuc_rt112_only <- setdiff(umuc_rt112_pparg_signature, genes_use_pparg)
both_umuc9_umuc_rt112   <- intersect(genes_use_pparg, umuc_rt112_only)

counts <- data.frame(
  region = c("UMUC9 only", "Both", "UMUC+RT112 only"),
  x = c(-0.8, 0, 0.8),
  y = c(0, 0, 0),
  label = c(length(umuc9_only), length(both_umuc9_umuc_rt112), length(umuc_rt112_only))
)

highlight_df <- data.frame(gene = highlight_genes) %>%
  filter(gene %in% union(umuc9_only, umuc_rt112_only)) %>%
  mutate(
    region = case_when(
      gene %in% both_umuc9_umuc_rt112 ~ "Both",
      gene %in% umuc9_only ~ "A only",
      gene %in% umuc_rt112_only ~ "B only"
    ),
    x = case_when(
      region == "A only" ~ -0.8,
      region == "Both" ~ 0,
      region == "B only" ~ 0.8
    ),
    y = row_number() * 0.15
  )


ggplot() +
  geom_circle(
    aes(x0 = -0.5, y0 = 0, r = 1),
    fill = "skyblue",
    alpha = 0.35,
    color = "black"
  ) +
  geom_circle(
    aes(x0 = 0.5, y0 = 0, r = 1),
    fill = "salmon",
    alpha = 0.35,
    color = "black"
  ) +
  geom_text(
    data = counts,
    aes(x = x, y = y, label = label),
    size = 7
  ) +
  geom_text_repel(
    data = highlight_df,
    aes(x = x, y = y, label = gene),
    size = 4,
    fontface = "bold",
    box.padding = 0.4
  ) +
  annotate("text", x = -0.9, y = 1.15, label = "UMUC9", size = 5) +
  annotate("text", x = 0.9, y = 1.15, label = "UMUC+RT112", size = 5) +
  coord_fixed() +
  theme_void()
```
