---
title: "run_base47_rnaseq"
output: html_document
---

bulk RNA-seq per-sample expression matrix
↓
normalized/log-transformed expression
↓
47 BASE47 genes only
↓
RNA-seq BASE47 classifier
↓
Basal vs Luminal call per sample

Install packages
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

if (!requireNamespace("DESeq2", quietly = TRUE)) {
  BiocManager::install("DESeq2")
}

if (!requireNamespace("biomaRt", quietly = TRUE)) {
  BiocManager::install("biomaRt")
}

if (!requireNamespace("pamr", quietly = TRUE)) {
  install.packages("pamr")
}

library(DESeq2)
library(pamr)
library(tidyverse)



```
Subset to BASE47 genes
```{r}
#BASE47 genes
base47_genes <- c(
  "ALOX5AP", "BHMT", "CAPN5", "CD14", "CDK6", "CYP2J2", "CYP4B1",
  "EMP3", "FAP", "FBP1", "GATA3", "HMGCS2", "MSN", "MT1X", "MT2A",
  "PRRX1", "PPARG", "SCNN1B", "SCNN1G", "SLC9A2", "SPINK1", "TBX2",
  "TMPRSS2", "UPK2", "PPFIBP2", "SEMA5A", "ADIRF", "SLC27A2",
  "GLIPR1", "UPK1A", "TRAK1", "PALLD", "GPD1L", "TOX3", "TMEM97",
  "CHST15", "VGLL1", "PLEKHG6", "PDGFC", "GAREM1", "GDPD3", "RNF128",
  "TUBB6", "CAVIN3", "AHNAK2", "RAB15", "FAM174B"
)
```

PREPARE RNAseq Matrix into BASE47 input format
- raw data for BBN+UPPL is cpm but BASE47 requires tpm
```{r}
#expression matrix load in
RNAseq_raw <- read.table("~/Documents/GitHub/EV-Resistance-analysis/BASE47/data/parental_mp1_K8HVZP-expression-matrix.csv", sep=",", row.names=1, header=T)

#Drop protein-coding identity + counts columns // keep only cpm columns
RNAseq_raw <- RNAseq_raw[, !grepl("count|^gene_biotype", colnames(RNAseq_raw)    )    ]

# Convert mouse ensembl id to human gene id (MANUALLY, b/c bioMarT was down)
base47_mouse_map <- read.table(
  text = "
ENSMUSG00000000093	TBX2
ENSMUSG00000000216	SCNN1G
ENSMUSG00000000385	TMPRSS2
ENSMUSG00000000392	FAP
ENSMUSG00000000440	PPARG
ENSMUSG00000001473	TUBB6
ENSMUSG00000006313	UPK1A
ENSMUSG00000015619	GATA3
ENSMUSG00000021062	RAB15
ENSMUSG00000022231	SEMA5A
ENSMUSG00000024503	SPINK1
ENSMUSG00000026062	SLC9A2
ENSMUSG00000026586	PRRX1
ENSMUSG00000027359	SLC27A2
ENSMUSG00000027875	HMGCS2
ENSMUSG00000028019	PDGFC
ENSMUSG00000028713	CYP4B1
ENSMUSG00000030703	GDPD3
ENSMUSG00000030873	SCNN1B
ENSMUSG00000030930	CHST15
ENSMUSG00000031131	VGLL1
ENSMUSG00000031207	MSN
ENSMUSG00000031438	RNF128
ENSMUSG00000032536	TRAK1
ENSMUSG00000035547	CAPN5
ENSMUSG00000036528	PPFIBP2
ENSMUSG00000037060	PRKCDBP
ENSMUSG00000037278	TMEM97
ENSMUSG00000038167	PLEKHG6
ENSMUSG00000040212	EMP3
ENSMUSG00000040274	CDK6
ENSMUSG00000041523	UPK2
ENSMUSG00000042680	GAREM
ENSMUSG00000043668	TOX3
ENSMUSG00000050627	GPD1L
ENSMUSG00000051439	CD14
ENSMUSG00000056888	GLIPR1
ENSMUSG00000058056	PALLD
ENSMUSG00000060063	ALOX5AP
ENSMUSG00000069805	FBP1
ENSMUSG00000074768	BHMT
ENSMUSG00000078670	FAM174B
ENSMUSG00000118667	AHNAK2
ENSMUSG00000031762	MT2A
ENSMUSG00000031765	MT1X
ENSMUSG00000052914	CYP2J2
NONE	ADIRF
",
  sep = "\t",
  col.names = c("mouse_ensembl", "human_gene"),
  stringsAsFactors = FALSE
)

# Convert mouse ensembl ID to human orthologs
RNAseq_raw_base47 <- RNAseq_raw %>%
  as.data.frame() %>%
  rownames_to_column("mouse_ensembl") %>%
  dplyr::mutate(mouse_ensembl_clean = sub("\\..*$", "", mouse_ensembl)) %>%
  dplyr::inner_join(
    base47_mouse_map %>%
      dplyr::filter(mouse_ensembl != "NONE") %>%
      dplyr::select(mouse_ensembl, human_gene),
    by = c("mouse_ensembl_clean" = "mouse_ensembl")
  ) %>%
  dplyr::select(-dplyr::any_of(c("mouse_ensembl", "mouse_ensembl_clean", "gene_type"))) %>%
  column_to_rownames("human_gene")
RNAseq_raw_base47$gene_name <- NULL

# mouse genes that didn't have human ortholog
missing_genes <- setdiff(
  base47_mouse_map$human_gene,
  rownames(RNAseq_raw_base47)
)
missing_genes

# Ensure matrix is numeric
RNAseq_raw_base47 <- as.data.frame(RNAseq_raw_base47)
RNAseq_raw_base47[] <- lapply(RNAseq_raw_base47, as.numeric)

# Convert cpm to log2 upper quartile-normalized expression (that's how they did it in their methods)
#   1. For each sample, calculate 75th percentile of nonzero CPM values
#   2. Divide each sample by its own upper-quartile value
#   3. Multiply by 1000 to put values on comparable scale
#   4. Apply log2(x+1) transformation
cpm_to_uq_log2 <- function(cpm_matrix) {
  uq <- apply(cpm_matrix, 2, function(x) quantile(x[x > 0], 0.75))
  scaled <- sweep(cpm_matrix, 2, uq, "/") * 1000
  log2(scaled + 1)
}

base47_uqlog2 <- cpm_to_uq_log2(RNAseq_raw_base47)

#centers genes across samples
base47_input <- t(scale(t(base47_uqlog2), center = TRUE, scale = FALSE))

#For each gene/row:
#   centered value = sample value - median expression of that gene across samples
base47_input <- sweep(base47_uqlog2, 1, apply(base47_uqlog2, 1, median, na.rm = TRUE), "-")

# Every row median should now be exactly 0 (sanity check that matrix is converted correctly)
stopifnot(all(abs(apply(base47_input, 1, median, na.rm = TRUE)) < 1e-10))


missing_genes <- setdiff(rownames(xn), rownames(base47_input))

#Checks for missing BASE47 genes in our data
# - it's ok if it stops the script, it's just mainly to let you know which ones are missing
# - if missing any genes, then run the BASE47 classifier on only the genes we have, rather than making the expression=0 for that missing gene
if (length(missing_genes) > 0) {
  stop("Missing BASE47 genes from df: ", paste(missing_genes, collapse = ", "))
}


```


Download BASE47 classifier model
```{r}
# Load BASE47 RNA-seq prediction matrix
download.file(
  "https://raw.githubusercontent.com/kimlabunc/Kardos_Rose_NanoString/main/BASE47_RNASEQ_PREDICTOR_DO_NOT_EDIT.txt",
  destfile = "BASE47_RNASEQ_PREDICTOR_DO_NOT_EDIT.txt"
)
x <- read.table("BASE47_RNASEQ_PREDICTOR_DO_NOT_EDIT.txt",sep="\t",row.names=1,header=T)

# gets labels for subtypes from 1ST ROW where: 1 = Basal; 2=Luminal
classes <- as.vector(t(x[1,]))

# removes 1st row, converts exp values to numeric, and restores gene names as row names
xn <- apply(x[-1,],2,function(x){as.numeric(as.vector(x))}) # xn is training expression matrix: genes by samples
rownames(xn) <- rownames(x)[-1]

# if missing any genes, remove it from prediction matrix
# rather than making expression=0
common_genes <- intersect(rownames(base47_uqlog2), rownames(xn))

# Optional: see what was removed
missing_from_sample <- setdiff(rownames(xn), rownames(base47_uqlog2))
missing_from_predictor <- setdiff(rownames(base47_uqlog2), rownames(xn))

missing_from_sample
missing_from_predictor

# Subset both matrices to the same genes, same order
base47_uqlog2_matched <- base47_uqlog2[common_genes, , drop = FALSE]
xn_matched <- xn[common_genes, , drop = FALSE]

# Classify BBN963+UPPL1541 RNAseq file
df <- base47_input
trainSet <- list(x=scale(xn_matched), y=classes, geneid=rownames(xn_matched), genenames=rownames(xn_matched))
mytrain <-pamr.train(trainSet)
pred.class <- pamr.predict(mytrain,scale(df),0)
pred.prob <- pamr.predict(mytrain,scale(df),0,type="posterior")
unc.tma.rna.calls <- as.data.frame(cbind(pred.prob,pred.class))
colnames(unc.tma.rna.calls) <- c("Basal.Correlation","Luminal.Correlation","BASE47.Subtype")
unc.tma.rna.calls$BASE47.Subtype[unc.tma.rna.calls$BASE47.Subtype == 1] <- "Basal"
unc.tma.rna.calls$BASE47.Subtype[unc.tma.rna.calls$BASE47.Subtype == 2] <- "Luminal"

write.csv(unc.tma.rna.calls, file="BASE47_classification_output.csv", row.names=TRUE)
```

Visual plot showing results

```{r}

#Create new columns for expected mol subtype and its BASE47 call
unc.tma.rna.calls <- unc.tma.rna.calls %>%
  rownames_to_column("Sample") %>%
  mutate(
    Tumor.Model = case_when(
      str_detect(Sample, "^BBN") ~ "BBN",
      str_detect(Sample, "^UPPL") ~ "UPPL",
      TRUE ~ NA_character_
    ),
    Basal.Score = Basal.Correlation - Luminal.Correlation
  ) %>%
  arrange(Basal.Score) %>%
  mutate(Sample = factor(Sample, levels = Sample))

# Confusion matrix
confusion_mtx <- table(
  Expected = unc.tma.rna.calls$Tumor.Model,
  Called = unc.tma.rna.calls$BASE47.Subtype
)
confusion_mtx <- as.data.frame.matrix(confusion_mtx)


# Waterfall plot
wfall_basal_score <- ggplot(unc.tma.rna.calls, aes(x=Sample, y=Basal.Score, color=Tumor.Model, fill=Tumor.Model)) + 
  geom_bar(stat = "identity") +
  scale_fill_hue(c = 40) +
  theme(legend.position="none") + theme_minimal() + coord_flip() + 
  labs(
  x = "Sample",
  y = "Basal centroid score: Luminal = -1, Basal = +1"
)
wfall_basal_score
ggsave("basal_score_waterfall.png", wfall_basal_score, width = 7, height = 5, dpi = 300)

```

GATA3 + PPARG Activity Scoring
```{r}
BiocManager::install("GEOquery")
BiocManager::install("impute")
library(samr)
library(GEOquery)
library(limma)

gse <- getGEO("GSE47993", GSEMatrix = TRUE)[[1]]
expr <- exprs(gse)
pdata <- pData(gse)

expr_umuc9 <- expr[, c("GSM1164283","GSM1164284","GSM1164285",  # control samples
           "GSM1164286","GSM1164287","GSM1164288") # rosi treated samples
           ]

# Get gene names from illumina probe IDs (they used IlluminaHT-12 V4.0 expression beadchip)
anno <- fData(gse)
gene_symbols <- anno[rownames(expr_umuc9), "Symbol"]

# keep probes with valid gene symbols
expr_annotated <- expr_umuc9[!is.na(gene_symbols) & gene_symbols != "", ]
gene_symbols <- gene_symbols[!is.na(gene_symbols) & gene_symbols != ""]
# Collapse multiple probes per gene
expr_gene <- avereps(expr_annotated, ID = gene_symbols) # now have gene expression x sample matrix

sam_data <- list(
  x = as.matrix(expr_gene),
  y = c(1,1,1,2,2,2), #1=ctrl, #2=rosi-treated
  geneid = rownames(expr_gene),
  genenames = rownames(expr_gene),
  logged2 = TRUE
)

samr_obj <- samr(sam_data, resp.type = "Two class unpaired", nperms = 100)
delta_table <- samr.compute.delta.table(samr_obj)

# Choose a delta where median FDR < 0.05
sig <- samr.compute.siggenes.table(
  samr.obj = samr_obj,
  del = chosen_delta,
  data = sam_data,
  delta.table = delta_table
)

pparg_sig <- sig$genes.up[, "Gene Name"]



```


